---
title: "Data cleaning"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
jupyter: R
---

## Setup

Load cleaning setup and project packages. Raw data lives in `data_raw/`; we use the NUTS boundaries GeoPackage.

```{r setup, include = FALSE}
source(here::here("data_cleaning.R"))
```

## NUTS boundaries (data_raw)

Read the NUTS regions from GISCO: **NUTS_RG_20M_2024_3035.gpkg** (polygons, 1:20M, EPSG:3035).

```{r read-nuts}
nuts <- read_nuts_raw()
```

### Structure

```{r nuts-structure}
# CRS and extent
st_crs(nuts)
st_bbox(nuts)

# Dimensions and column names
dim(nuts)
names(nuts)
```

### Preview

```{r nuts-preview}
# First rows (geometry omitted for display)
nuts |> 
  st_drop_geometry() |> 
  head(10)
```

### NUTS levels

```{r nuts-levels}
# Typical GISCO attributes: LEVL_CODE (0=country, 1–3=NUTS1–3), NUTS_ID, NAME_LATN, etc.
if ("LEVL_CODE" %in% names(nuts)) {
  nuts |> 
    st_drop_geometry() |> 
    count(LEVL_CODE, name = "n_regions")
}
```

### Simple map (NUTS)

```{r map-nuts, fig.width = 8, fig.height = 6}
library(tmap)
tm_shape(nuts) +
  tm_polygons(col = "grey85", border.col = "grey50") +
  tm_layout(title = "NUTS (2024, 20M)", frame = FALSE)
```

## Next steps

- Filter to study area or NUTS level (e.g. NUTS 1).
- Reproject to WGS84 (EPSG:4326) if needed for E-OBS/crop rasters; save to `data_proc/`.
