---
title: "Data cleaning"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
jupyter: R
---

## Setup

Load cleaning setup and project packages. Raw data lives in `data_raw/`; we use the NUTS boundaries GeoPackage.

```{r setup, include = FALSE}
source(here::here("data_cleaning.R"))
```

```{r check-terra}
Sys.setenv(PROJ_LIB = system.file("proj", package = "terra"))

library(terra)
```

```{r check-proj}
system.file("proj", package = "terra")
# should be something like ".../R/win-library/4.5/terra/proj"
```

## Step 1 — Load everything and check metadata

### 1.1 Load libraries

```{r load-libs}
library(terra)
library(sf)
library(exactextractr)
library(sfdep)
library(dplyr)
library(tidyr)
library(tmap)
library(here)
```

### 1.2 Read datasets

- E-OBS TG NetCDF (daily)
- E-OBS RR NetCDF (daily)
- SPAM crop GeoTIFF(s)
- Admin boundaries (gpkg/shp)

```{r read-datasets}
# E-OBS temperature (TG) and precipitation (RR)
tg <- rast(here("data_raw/eobs/tg_ens_mean_0.1deg_reg_2011-2022_v27.0e.nc"))
rr <- rast(here("data_raw/eobs/rr_ens_mean_0.1deg_reg_2011-2022_v27.0e.nc"))

# Admin boundaries
admin <- st_read(here("data_raw/admin/NUTS_RG_20M_2024_3035.gpkg"), quiet = TRUE)

# SPAM crop data (adjust path/pattern as needed)
# spam_files <- list.files(here("data_raw/spam"), pattern = "\\.tif$", full.names = TRUE)
# spam <- rast(spam_files)  # or read individual files
```

### 1.3 Inspect CRS + resolution + time dimension

```{r inspect-metadata}
crs(tg); res(tg); nlyr(tg)
crs(rr); res(rr); nlyr(rr)
st_crs(admin)
```

**What this answers:** "Do my inputs align technically?"

**Output:** a short "data QA" note in your report.

## Step 2 — Harmonize study area (DE–BE–NL) + fix geometries

### 2.1 Filter Admin polygons to DE/BE/NL

(Your attribute names differ; adjust `CNTR_CODE` / `ISO` / etc.)

```{r filter-debenl}
admin_debenl<- admin%>%
  filter(CNTR_CODE%in%c("DE","BE","NL"))
```

### 2.2 Fix geometry + create mask

```{r fix-geometry-mask}
admin_debenl<- st_make_valid(admin_debenl)
mask_union_4326<- st_union(st_transform(admin_debenl,4326))
mask_v<- vect(mask_union_4326)
```

## Step 3 — Clip E-OBS rasters to DE–BE–NL

```{r clip-eobs}
tg_clip<- mask(crop(tg, mask_v), mask_v)
rr_clip<- mask(crop(rr, mask_v), mask_v)
```

**Outputs saved:**

```{r save-clipped-rasters}
dir.create(here("data_proc", "rasters"), showWarnings = FALSE, recursive = TRUE)
writeRaster(tg_clip, here("data_proc/rasters/tg_debenl.tif"), overwrite=TRUE)
writeRaster(rr_clip, here("data_proc/rasters/rr_debenl.tif"), overwrite=TRUE)
```

## Step 4 — Select the time window (April–May)

E-OBS NetCDF usually has a time index. Two robust approaches:

### 4.1 If time labels exist (best)

```{r check-time}
time(tg_clip)  # check if terra shows dates
```

If dates exist, filter by month:

```{r filter-april-may}
dates<- as.Date(time(tg_clip))
sel<- which(format(dates,"%m")%in%c("04","05"))

tg_am<- tg_clip[[sel]]
rr_am<- rr_clip[[sel]]
```

## Step 5 — Compute daily suitability (PSI) from E-OBS

Define rules (example you already had):

- temp suitable: 12–25°C
- rain suitable: RR < 1 mm/day

### 5.1 Binary suitability per day

```{r psi-daily}
temp_ok<- tg_am>=12& tg_am<=25
rain_ok<- rr_am<1

psi_daily<- temp_ok* rain_ok  # 1 if both true else 0
```

### 5.2 Save daily PSI (optional but nice)

```{r save-psi-daily}
writeRaster(psi_daily, here("data_proc/rasters/psi_daily.tif"), overwrite=TRUE)
```

**Outputs you'll later summarize:** mean, count, variability.

## Step 6 — Temporal summaries (this is the "spatio-temporal" core)

You should compute at least these 3:

### 6.1 Mean suitability (fraction of suitable days)

```{r psi-mean}
psi_mean<- app(psi_daily, mean, na.rm=TRUE)  # 0–1
```

### 6.2 Number of suitable days

```{r psi-days}
psi_days<- app(psi_daily,sum, na.rm=TRUE)  # count of days
```

### 6.3 Variability / instability (SD)

```{r psi-sd}
psi_sd<- app(psi_daily, sd, na.rm=TRUE)
```

Save them:

```{r save-psi-summaries}
writeRaster(psi_mean, here("data_proc/rasters/psi_mean_am.tif"), overwrite=TRUE)
writeRaster(psi_days, here("data_proc/rasters/psi_days_am.tif"), overwrite=TRUE)
writeRaster(psi_sd,   here("data_proc/rasters/psi_sd_am.tif"), overwrite=TRUE)
```

```{r check-rasters}
library(terra)

psi_mean_am <- rast(here("data_proc/rasters/psi_mean_am.tif"))
psi_days_am <- rast(here("data_proc/rasters/psi_days_am.tif"))
psi_sd_am   <- rast(here("data_proc/rasters/psi_sd_am.tif"))
tg_debenl   <- rast(here("data_proc/rasters/tg_debenl.tif"))
rr_debenl   <- rast(here("data_proc/rasters/rr_debenl.tif"))
psi_daily   <- rast(here("data_proc/rasters/psi_daily.tif"))   # if you saved it too
```

```{r check}
# Are rasters empty?
#global(psi_mean_am, "mean", na.rm=TRUE)
#global(psi_days_am, "max", na.rm=TRUE)

# Value ranges
#global(tg_debenl, range, na.rm=TRUE)
#global(rr_debenl, range, na.rm=TRUE)

# PSI logic check
unique(values(psi_daily)[!is.na(values(psi_daily))])
```

## Step 7 — Bring SPAM 2020 crop data into the same grid

This is where SPAM shines: it's already ~10 km, but it may still differ from E-OBS grid.

### 7.1 Load SPAM crop raster (example: harvested area for rapeseed)

```{r spam-load}
spam_rape_ha<- rast(here("data_raw/spam/spam2020_V2r0_global_H_RAPE_A.tif"))
```

### 7.2 Normalize CRS and resample SPAM to E-OBS template

Use `psi_mean` as the template because it's exactly your analysis grid.

- Here we **force both rasters to EPSG:4326** and then use `resample()`. This avoids PROJ/GDAL transformation issues.
- If SPAM is continuous (ha, production): use `"bilinear"` or `"average"` style resampling.

```{r spam-project}
crs(spam_rape_ha)<- "EPSG:4326"
crs(psi_mean)    <- "EPSG:4326"

spam_rape_ha_eobs<- resample(spam_rape_ha, psi_mean, method="bilinear")
```

Save:

```{r spam-save}
writeRaster(spam_rape_ha_eobs, here("data_proc/rasters/spam_rape_ha_eobs.tif"), overwrite=TRUE)
```

## Step 8 — Convert SPAM “harvested area” to a crop fraction/intensity (optional but strong)

### 8.1 Compute cell area (ha) for the E-OBS grid

```{r cell-area}
cell_area_m2<- cellSize(psi_mean, unit="m")  # raster of m²
cell_area_ha<- cell_area_m2/10000
```

### 8.2 Crop fraction (0–1-ish, can exceed 1 in rare edge cases; clamp)

```{r rape-fraction}
rape_frac<- spam_rape_ha_eobs/ cell_area_ha
rape_frac<- clamp(rape_frac,0,1)
writeRaster(rape_frac, here("data_proc/rasters/rape_fraction.tif"), overwrite=TRUE)
```

## Step 9 — Compute crop-weighted suitability and crop-weighted risk

You want metrics that combine climate + crop exposure.

### 9.1 Crop-weighted suitability

```{r rape-suitability-weighted}
rape_suit<- psi_mean* rape_frac
writeRaster(rape_suit, here("data_proc/rasters/rape_suitability_weighted.tif"), overwrite=TRUE)
```

### 9.2 Crop-weighted “bad days” (exposure to unsuitable days)

First define “bad day” = PSI == 0:

```{r rape-bad-exposure}
bad_daily<- 1- psi_daily
bad_days<- app(bad_daily,sum, na.rm=TRUE)  # count of bad days

rape_bad_exposure<- bad_days* rape_frac
writeRaster(rape_bad_exposure, here("data_proc/rasters/rape_bad_exposure.tif"), overwrite=TRUE)
```

## Step 10 — Aggregate rasters to Admin regions


### 10.1 Prep admin polygons in same CRS as rasters

```{r admin-4326}
admin_4326<- st_transform(admin_debenl, crs(psi_mean))
```

### 10.2 Zonal stats with exactextractr

```{r admin-stats}
admin_stats<- admin_4326%>%
  mutate(
    psi_mean = exact_extract(psi_mean,           ., "mean"),
    psi_days = exact_extract(psi_days,           ., "mean"),
    psi_sd   = exact_extract(psi_sd,             ., "mean"),
    rape_frac= exact_extract(rape_frac,          ., "mean"),
    rape_suit= exact_extract(rape_suit,          ., "mean"),
    rape_bad = exact_extract(rape_bad_exposure,  ., "mean")
  )
```

Save table:

```{r admin-stats-save}
write.csv(
  st_drop_geometry(admin_stats),
  here("data_proc/tables/admin_pollination_stats.csv"),
  row.names = FALSE
)
```

**Outputs:**

- A clean regional dataset to map + analyze + rank.

## Step 11 — Spatial weights + Global Moran’s I + LISA (region level)

### 11.1 Prepare neighbors (queen contiguity)

```{r moran-neighbors}
admin_stats_3035<- st_transform(admin_stats, 3035)

nb<- st_contiguity(admin_stats_3035)      # neighbors (queen contiguity)
lw<- st_weights(nb, style="W")           # row-standardized weights
```

### 11.2 Global Moran’s I (example on rape_suit)
```{r}
summary(admin_stats_3035$rape_suit)
table(is.finite(admin_stats_3035$rape_suit))
which(!is.finite(admin_stats_3035$rape_suit))[1:20]  # first few indices
```
```{r global-moran}
library(dplyr)
library(sfdep)

# 1) Keep only regions with valid values
admin_ok <- admin_stats_3035 %>%
  filter(!is.na(rape_suit), is.finite(rape_suit))

# report how many were removed
cat("Rows total:", nrow(admin_stats_3035),
    " | used:", nrow(admin_ok),
    " | removed:", nrow(admin_stats_3035) - nrow(admin_ok), "\n")

# 2) Build neighbors + weights (queen contiguity)
nb_ok <- st_contiguity(admin_ok)
zero_nbr <- lengths(nb_ok) == 0
if (any(zero_nbr)) {
  cat("Dropping isolated regions (0 neighbors):", sum(zero_nbr), "\n")
  admin_ok <- admin_ok[!zero_nbr, ]
  nb_ok <- st_contiguity(admin_ok)
}
lw_ok <- st_weights(nb_ok, style = "W")

# 3) Global Moran's I (sfdep versions without allow_zero argument)
global_m <- global_moran(admin_ok$rape_suit, nb_ok, lw_ok)
global_m

# (optional) permutation-based p-value (stronger for reporting)
global_mc <- global_moran_perm(admin_ok$rape_suit, nb_ok, lw_ok, nsim = 999)
global_mc

```

### 11.3 Local Moran’s I (LISA)

```{r local-moran}
library(dplyr)
library(sfdep)

# 0) Create a stable ID (do this ONCE)
admin_stats_3035 <- admin_stats_3035 %>%
  mutate(region_id = row_number())

# 1) Keep only finite values
admin_ok <- admin_stats_3035 %>%
  filter(is.finite(rape_suit))

# 2) Rebuild neighbors/weights
nb_ok <- st_contiguity(admin_ok)
zero_nbr_lisa <- lengths(nb_ok) == 0
if (any(zero_nbr_lisa)) {
  cat("Dropping isolated regions for LISA:", sum(zero_nbr_lisa), "\n")
  admin_ok <- admin_ok[!zero_nbr_lisa, ]
  nb_ok <- st_contiguity(admin_ok)
}
lw_ok <- st_weights(nb_ok, style = "W")

# 3) LISA
lisa_ok <- local_moran(admin_ok$rape_suit, nb_ok, lw_ok)

# 4) Attach results (robust across sfdep output variants)
admin_ok <- admin_ok %>%
  bind_cols(as_tibble(lisa_ok)) %>%
  mutate(
    lisa_i  = ii,
    lisa_p  = p_ii
  )

if ("mean" %in% names(admin_ok)) {
  admin_ok <- admin_ok %>% mutate(lisa_cl = as.character(mean))
} else if ("cluster" %in% names(admin_ok)) {
  admin_ok <- admin_ok %>% mutate(lisa_cl = as.character(cluster))
} else {
  admin_ok <- admin_ok %>% mutate(lisa_cl = NA_character_)
}


```

```{r lisa-join}
# 5) Join back using region_id
admin_stats_3035 <- admin_stats_3035 %>%
  left_join(
    admin_ok %>%
      st_drop_geometry() %>%
      select(region_id, lisa_i, lisa_p, lisa_cl),
    by = "region_id"
  )
```

Verify the join:

```{r lisa-verify}
# Should show NAs for regions dropped from admin_ok (non-finite rape_suit) and the rest filled
table(is.na(admin_stats_3035$lisa_i))
table(admin_stats_3035$lisa_cl, useNA = "ifany")
summary(admin_stats_3035$lisa_p)
```

Create a "significant cluster" label (for mapping):

```{r lisa-sig}
admin_stats_3035 <- admin_stats_3035 %>%
  mutate(
    lisa_sig = case_when(
      is.na(lisa_p) ~ NA_character_,
      lisa_p <= 0.05 ~ as.character(lisa_cl),
      TRUE ~ "Not significant"
    )
  )

table(admin_stats_3035$lisa_sig, useNA = "ifany")
```


Save:

```{r lisa-save}
st_write(
  admin_stats_3035,
  here("data_proc/vectors/admin_stats_lisa.gpkg"),
  delete_dsn = TRUE
)
```

## Step 12 - Create final deliverables (maps, plots, tables)

### 12.1 Maps (minimum set)

```{r maps-deliverables}
library(tmap)
library(here)

dir.create(here("outputs", "maps"), showWarnings = FALSE, recursive = TRUE)
tmap_mode("plot")

save_map <- function(map_obj, filename) {
  tmap_save(map_obj, filename = filename, width = 10, height = 7)
}

map_psi_mean <- tm_shape(psi_mean) +
  tm_raster(title = "PSI mean") +
  tm_layout(title = "Mean PSI (Apr-May)", frame = FALSE)

map_psi_days <- tm_shape(psi_days) +
  tm_raster(title = "Suitable days") +
  tm_layout(title = "Suitable days count (Apr-May)", frame = FALSE)

map_psi_sd <- tm_shape(psi_sd) +
  tm_raster(title = "PSI SD") +
  tm_layout(title = "PSI variability (SD, Apr-May)", frame = FALSE)

map_rape_frac <- tm_shape(rape_frac) +
  tm_raster(title = "Rape fraction") +
  tm_layout(title = "Rapeseed fraction/intensity", frame = FALSE)

map_rape_suit <- tm_shape(rape_suit) +
  tm_raster(title = "Weighted suitability") +
  tm_layout(title = "Crop-weighted suitability", frame = FALSE)

save_map(map_psi_mean, here("outputs/maps/map_psi_mean.png"))
save_map(map_psi_days, here("outputs/maps/map_psi_days.png"))
save_map(map_psi_sd, here("outputs/maps/map_psi_sd.png"))
save_map(map_rape_frac, here("outputs/maps/map_rape_frac.png"))
save_map(map_rape_suit, here("outputs/maps/map_rape_suit.png"))

if ("lisa_sig" %in% names(admin_stats_3035)) {
  map_lisa <- tm_shape(admin_stats_3035) +
    tm_polygons("lisa_sig", title = "LISA cluster", colorNA = "grey85") +
    tm_layout(title = "LISA clusters (rape_suit)", frame = FALSE)
  save_map(map_lisa, here("outputs/maps/map_lisa_clusters.png"))
}

map_psi_mean
```

### 12.2 Time series plot per region (region x day)

```{r region-day-timeseries}
library(dplyr)
library(tidyr)
library(exactextractr)
library(ggplot2)
library(here)

dir.create(here("outputs", "plots"), showWarnings = FALSE, recursive = TRUE)
dir.create(here("data_proc", "tables"), showWarnings = FALSE, recursive = TRUE)

admin_ts <- st_transform(admin_debenl, crs(psi_daily))
admin_labels <- st_drop_geometry(admin_ts)
admin_labels$region_id <- seq_len(nrow(admin_labels))

label_candidates <- c("NUTS_ID", "NAME_LATN", "NUTS_NAME", "NAME_ENGL")
label_col <- label_candidates[label_candidates %in% names(admin_labels)]
if (length(label_col) > 0) {
  admin_labels$region_label <- as.character(admin_labels[[label_col[1]]])
} else {
  admin_labels$region_label <- paste0("region_", admin_labels$region_id)
}

daily_wide <- exact_extract(psi_daily, admin_ts, "mean") %>% as_tibble()
daily_wide$region_id <- seq_len(nrow(daily_wide))

layer_lookup <- tibble(
  layer = names(psi_daily),
  layer_index = seq_len(nlyr(psi_daily))
)

psi_dates <- suppressWarnings(as.Date(time(psi_daily)))
if (length(psi_dates) != nlyr(psi_daily)) {
  psi_dates <- rep(as.Date(NA), nlyr(psi_daily))
}
if (all(is.na(psi_dates)) && exists("tg_dates") && exists("sel") && length(sel) == nlyr(psi_daily)) {
  psi_dates <- tg_dates[sel]
}
layer_lookup$date <- psi_dates

psi_region_day <- daily_wide %>%
  pivot_longer(-region_id, names_to = "layer", values_to = "psi_daily_mean") %>%
  left_join(layer_lookup, by = "layer") %>%
  left_join(admin_labels %>% select(region_id, region_label), by = "region_id")

write.csv(psi_region_day, here("data_proc/tables/psi_region_day.csv"), row.names = FALSE)

if (!("region_id" %in% names(admin_stats_3035))) {
  admin_stats_3035 <- admin_stats_3035 %>% mutate(region_id = row_number())
}

top_regions <- admin_stats_3035 %>%
  st_drop_geometry() %>%
  filter(is.finite(rape_bad)) %>%
  arrange(desc(rape_bad)) %>%
  slice_head(n = 12) %>%
  pull(region_id)

plot_df <- psi_region_day %>%
  filter(region_id %in% top_regions)

if (all(is.na(plot_df$date))) {
  p_ts <- ggplot(plot_df, aes(x = layer_index, y = psi_daily_mean, color = region_label, group = region_label)) +
    geom_line(alpha = 0.8) +
    labs(
      x = "Day index (Apr-May)",
      y = "Daily regional mean PSI",
      title = "Daily PSI by region (top 12 by rape_bad)"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
} else {
  p_ts <- ggplot(plot_df, aes(x = date, y = psi_daily_mean, color = region_label, group = region_label)) +
    geom_line(alpha = 0.8) +
    labs(
      x = "Date",
      y = "Daily regional mean PSI",
      title = "Daily PSI by region (top 12 by rape_bad)"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

ggsave(here("outputs/plots/ts_psi_daily_top12_rape_bad.png"), p_ts, width = 10, height = 6, dpi = 300)
p_ts
```

### 12.3 Ranking tables (policy relevance)

```{r ranking-tables}
library(dplyr)
library(here)

dir.create(here("outputs", "tables"), showWarnings = FALSE, recursive = TRUE)

stats_tbl <- read.csv(here("data_proc/tables/admin_pollination_stats.csv"), stringsAsFactors = FALSE)
id_candidates <- c("NUTS_ID", "NAME_LATN", "CNTR_CODE", "LEVL_CODE")
id_cols <- id_candidates[id_candidates %in% names(stats_tbl)]

top_rape_bad <- stats_tbl %>%
  filter(is.finite(rape_bad)) %>%
  arrange(desc(rape_bad)) %>%
  slice_head(n = 10)

top_rape_suit <- stats_tbl %>%
  filter(is.finite(rape_suit)) %>%
  arrange(desc(rape_suit)) %>%
  slice_head(n = 10)

top_psi_sd <- stats_tbl %>%
  filter(is.finite(psi_sd)) %>%
  arrange(desc(psi_sd)) %>%
  slice_head(n = 10)

write.csv(top_rape_bad, here("outputs/tables/top10_rape_bad.csv"), row.names = FALSE)
write.csv(top_rape_suit, here("outputs/tables/top10_rape_suit.csv"), row.names = FALSE)
write.csv(top_psi_sd, here("outputs/tables/top10_psi_sd.csv"), row.names = FALSE)

top_rape_bad %>% select(any_of(c(id_cols, "rape_bad")))
top_rape_suit %>% select(any_of(c(id_cols, "rape_suit")))
top_psi_sd %>% select(any_of(c(id_cols, "psi_sd")))
```

## NUTS boundaries (data_raw)

Read the NUTS regions from GISCO: **NUTS_RG_20M_2024_3035.gpkg** (polygons, 1:20M, EPSG:3035).

```{r read-nuts}
nuts <- read_nuts_raw()
```

### Structure

```{r nuts-structure}
# CRS and extent
st_crs(nuts)
st_bbox(nuts)

# Dimensions and column names
dim(nuts)
names(nuts)
```

### Preview

```{r nuts-preview}
# First rows (geometry omitted for display)
nuts |> 
  st_drop_geometry() |> 
  head(10)
```

### NUTS levels

```{r nuts-levels}
# Typical GISCO attributes: LEVL_CODE (0=country, 1–3=NUTS1–3), NUTS_ID, NAME_LATN, etc.
if ("LEVL_CODE" %in% names(nuts)) {
  nuts |> 
    st_drop_geometry() |> 
    count(LEVL_CODE, name = "n_regions")
}
```

### Simple map (NUTS)

```{r map-nuts, fig.width = 8, fig.height = 6}
library(tmap)
tm_shape(nuts) +
  tm_polygons(col = "grey85", border.col = "grey50") +
  tm_layout(title = "NUTS (2024, 20M)", frame = FALSE)
```

## Next steps

- Filter to study area or NUTS level (e.g. NUTS 1).
- Reproject to WGS84 (EPSG:4326) if needed for E-OBS/crop rasters; save to `data_proc/`.
