---
title: "Data cleaning"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
jupyter: R
---

## Setup

Load cleaning setup and project packages. Raw data lives in `data_raw/`; we use the NUTS boundaries GeoPackage.

```{r setup, include = FALSE}
source(here::here("data_cleaning.R"))
```

```{r check-terra}
Sys.setenv(PROJ_LIB = system.file("proj", package = "terra"))

library(terra)
```

```{r check-proj}
system.file("proj", package = "terra")
# should be something like ".../R/win-library/4.5/terra/proj"
```

## Step 1 — Load everything and check metadata

### 1.1 Load libraries

```{r load-libs}
library(terra)
library(sf)
library(exactextractr)
library(sfdep)
library(dplyr)
library(tidyr)
library(tmap)
library(here)
```

### 1.2 Read datasets

- E-OBS TG NetCDF (daily)
- E-OBS RR NetCDF (daily)
- SPAM crop GeoTIFF(s)
- Admin boundaries (gpkg/shp)

```{r read-datasets}
# E-OBS temperature (TG) and precipitation (RR)
tg <- rast(here("data_raw/eobs/tg_ens_mean_0.1deg_reg_2011-2022_v27.0e.nc"))
rr <- rast(here("data_raw/eobs/rr_ens_mean_0.1deg_reg_2011-2022_v27.0e.nc"))

# Admin boundaries
admin <- st_read(here("data_raw/admin/NUTS_RG_20M_2024_3035.gpkg"), quiet = TRUE)

# SPAM crop data (adjust path/pattern as needed)
# spam_files <- list.files(here("data_raw/spam"), pattern = "\\.tif$", full.names = TRUE)
# spam <- rast(spam_files)  # or read individual files
```

### 1.3 Inspect CRS + resolution + time dimension

```{r inspect-metadata}
crs(tg); res(tg); nlyr(tg)
crs(rr); res(rr); nlyr(rr)
st_crs(admin)
```

**What this answers:** "Do my inputs align technically?"

**Output:** a short "data QA" note in your report.

## Step 2 — Harmonize study area (DE–BE–NL) + fix geometries

### 2.1 Filter Admin polygons to DE/BE/NL

(Your attribute names differ; adjust `CNTR_CODE` / `ISO` / etc.)

```{r filter-debenl}
admin_debenl<- admin%>%
  filter(CNTR_CODE%in%c("DE","BE","NL"))
```

### 2.2 Fix geometry + create mask

```{r fix-geometry-mask}
admin_debenl<- st_make_valid(admin_debenl)
mask_union_4326<- st_union(st_transform(admin_debenl,4326))
mask_v<- vect(mask_union_4326)
```

## Step 3 — Clip E-OBS rasters to DE–BE–NL

```{r clip-eobs}
tg_clip<- mask(crop(tg, mask_v), mask_v)
rr_clip<- mask(crop(rr, mask_v), mask_v)
```

**Outputs saved:**

```{r save-clipped-rasters}
dir.create(here("data_proc", "rasters"), showWarnings = FALSE, recursive = TRUE)
writeRaster(tg_clip, here("data_proc/rasters/tg_debenl.tif"), overwrite=TRUE)
writeRaster(rr_clip, here("data_proc/rasters/rr_debenl.tif"), overwrite=TRUE)
```

## Step 4 — Select the time window (April–May)

E-OBS NetCDF usually has a time index. Two robust approaches:

### 4.1 If time labels exist (best)

```{r check-time}
time(tg_clip)  # check if terra shows dates
```

If dates exist, filter by month:

```{r filter-april-may}
dates<- as.Date(time(tg_clip))
sel<- which(format(dates,"%m")%in%c("04","05"))

tg_am<- tg_clip[[sel]]
rr_am<- rr_clip[[sel]]
```

## Step 5 — Compute daily suitability (PSI) from E-OBS

Define rules (example you already had):

- temp suitable: 12–25°C
- rain suitable: RR < 1 mm/day

### 5.1 Binary suitability per day

```{r psi-daily}
temp_ok<- tg_am>=12& tg_am<=25
rain_ok<- rr_am<1

psi_daily<- temp_ok* rain_ok  # 1 if both true else 0
```

### 5.2 Save daily PSI (optional but nice)

```{r save-psi-daily}
writeRaster(psi_daily, here("data_proc/rasters/psi_daily.tif"), overwrite=TRUE)
```

**Outputs you'll later summarize:** mean, count, variability.

## Step 6 — Temporal summaries (this is the "spatio-temporal" core)

You should compute at least these 3:

### 6.1 Mean suitability (fraction of suitable days)

```{r psi-mean}
psi_mean<- app(psi_daily, mean, na.rm=TRUE)  # 0–1
```

### 6.2 Number of suitable days

```{r psi-days}
psi_days<- app(psi_daily,sum, na.rm=TRUE)  # count of days
```

### 6.3 Variability / instability (SD)

```{r psi-sd}
psi_sd<- app(psi_daily, sd, na.rm=TRUE)
```

Save them:

```{r save-psi-summaries}
writeRaster(psi_mean, here("data_proc/rasters/psi_mean_am.tif"), overwrite=TRUE)
writeRaster(psi_days, here("data_proc/rasters/psi_days_am.tif"), overwrite=TRUE)
writeRaster(psi_sd,   here("data_proc/rasters/psi_sd_am.tif"), overwrite=TRUE)
```

```{r check-rasters}
library(terra)

psi_mean_am <- rast(here("data_proc/rasters/psi_mean_am.tif"))
psi_days_am <- rast(here("data_proc/rasters/psi_days_am.tif"))
psi_sd_am   <- rast(here("data_proc/rasters/psi_sd_am.tif"))
tg_debenl   <- rast(here("data_proc/rasters/tg_debenl.tif"))
rr_debenl   <- rast(here("data_proc/rasters/rr_debenl.tif"))
psi_daily   <- rast(here("data_proc/rasters/psi_daily.tif"))   # if you saved it too
```

```{r check}
# Are rasters empty?
global(psi_mean_am, "mean", na.rm=TRUE)
global(psi_days_am, "max", na.rm=TRUE)

# Value ranges
global(tg_debenl, range, na.rm=TRUE)
global(rr_debenl, range, na.rm=TRUE)

# PSI logic check
unique(values(psi_daily)[!is.na(values(psi_daily))])
```

## Step 7 — Bring SPAM 2020 crop data into the same grid

This is where SPAM shines: it's already ~10 km, but it may still differ from E-OBS grid.

### 7.1 Load SPAM crop raster (example: harvested area for rapeseed)

```{r spam-load}
spam_rape_ha<- rast(here("data_raw/spam/spam2020_V2r0_global_H_RAPE_A.tif"))
```

### 7.2 Normalize CRS and resample SPAM to E-OBS template

Use `psi_mean` as the template because it's exactly your analysis grid.

- Here we **force both rasters to EPSG:4326** and then use `resample()`. This avoids PROJ/GDAL transformation issues.
- If SPAM is continuous (ha, production): use `"bilinear"` or `"average"` style resampling.

```{r spam-project}
crs(spam_rape_ha)<- "EPSG:4326"
crs(psi_mean)    <- "EPSG:4326"

spam_rape_ha_eobs<- resample(spam_rape_ha, psi_mean, method="bilinear")
```

Save:

```{r spam-save}
writeRaster(spam_rape_ha_eobs, here("data_proc/rasters/spam_rape_ha_eobs.tif"), overwrite=TRUE)
```

## Step 8 — Convert SPAM “harvested area” to a crop fraction/intensity (optional but strong)

### 8.1 Compute cell area (ha) for the E-OBS grid

```{r cell-area}
cell_area_m2<- cellSize(psi_mean, unit="m")  # raster of m²
cell_area_ha<- cell_area_m2/10000
```

### 8.2 Crop fraction (0–1-ish, can exceed 1 in rare edge cases; clamp)

```{r rape-fraction}
rape_frac<- spam_rape_ha_eobs/ cell_area_ha
rape_frac<- clamp(rape_frac,0,1)
writeRaster(rape_frac, here("data_proc/rasters/rape_fraction.tif"), overwrite=TRUE)
```

## Step 9 — Compute crop-weighted suitability and crop-weighted risk

You want metrics that combine climate + crop exposure.

### 9.1 Crop-weighted suitability

```{r rape-suitability-weighted}
rape_suit<- psi_mean* rape_frac
writeRaster(rape_suit, here("data_proc/rasters/rape_suitability_weighted.tif"), overwrite=TRUE)
```

### 9.2 Crop-weighted “bad days” (exposure to unsuitable days)

First define “bad day” = PSI == 0:

```{r rape-bad-exposure}
bad_daily<- 1- psi_daily
bad_days<- app(bad_daily,sum, na.rm=TRUE)  # count of bad days

rape_bad_exposure<- bad_days* rape_frac
writeRaster(rape_bad_exposure, here("data_proc/rasters/rape_bad_exposure.tif"), overwrite=TRUE)
```

## NUTS boundaries (data_raw)

Read the NUTS regions from GISCO: **NUTS_RG_20M_2024_3035.gpkg** (polygons, 1:20M, EPSG:3035).

```{r read-nuts}
nuts <- read_nuts_raw()
```

### Structure

```{r nuts-structure}
# CRS and extent
st_crs(nuts)
st_bbox(nuts)

# Dimensions and column names
dim(nuts)
names(nuts)
```

### Preview

```{r nuts-preview}
# First rows (geometry omitted for display)
nuts |> 
  st_drop_geometry() |> 
  head(10)
```

### NUTS levels

```{r nuts-levels}
# Typical GISCO attributes: LEVL_CODE (0=country, 1–3=NUTS1–3), NUTS_ID, NAME_LATN, etc.
if ("LEVL_CODE" %in% names(nuts)) {
  nuts |> 
    st_drop_geometry() |> 
    count(LEVL_CODE, name = "n_regions")
}
```

### Simple map (NUTS)

```{r map-nuts, fig.width = 8, fig.height = 6}
library(tmap)
tm_shape(nuts) +
  tm_polygons(col = "grey85", border.col = "grey50") +
  tm_layout(title = "NUTS (2024, 20M)", frame = FALSE)
```

## Next steps

- Filter to study area or NUTS level (e.g. NUTS 1).
- Reproject to WGS84 (EPSG:4326) if needed for E-OBS/crop rasters; save to `data_proc/`.
