---
title: "Data cleaning"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
jupyter: R
---

## Setup

Load cleaning setup and project packages. Raw data lives in `data_raw/`; we use the NUTS boundaries GeoPackage.

```{r setup, include = FALSE}
source(here::here("data_cleaning.R"))
```

## Load everything and check metadata

### 1.1 Load libraries

```{r load-libs}
library(terra)
library(sf)
library(exactextractr)
library(sfdep)
library(dplyr)
library(tidyr)
library(tmap)
library(here)
```

### 1.2 Read datasets

- E-OBS TG NetCDF (daily)
- E-OBS RR NetCDF (daily)
- SPAM crop GeoTIFF(s)
- Admin boundaries (gpkg/shp)

```{r read-datasets}
# E-OBS temperature (TG) and precipitation (RR)
tg <- rast(here("data_raw/eobs/tg_ens_mean_0.1deg_reg_2011-2022_v27.0e.nc"))
rr <- rast(here("data_raw/eobs/rr_ens_mean_0.1deg_reg_2011-2022_v27.0e.nc"))

# Admin boundaries
admin <- st_read(here("data_raw/admin/rr_ens_mean_0.1deg_reg_2011-2022_v27.0e.gpkg"), quiet = TRUE)

# SPAM crop data (adjust path/pattern as needed)
# spam_files <- list.files(here("data_raw/spam"), pattern = "\\.tif$", full.names = TRUE)
# spam <- rast(spam_files)  # or read individual files
```

### 1.3 Inspect CRS + resolution + time dimension

```{r inspect-metadata}
# E-OBS TG
cat("=== E-OBS TG (Temperature) ===\n")
cat("CRS:", crs(tg), "\n")
cat("Resolution:", paste(res(tg), collapse = " x "), "degrees\n")
cat("Layers (time steps):", nlyr(tg), "\n")
cat("Extent:", paste(ext(tg)[], collapse = ", "), "\n\n")

# E-OBS RR
cat("=== E-OBS RR (Precipitation) ===\n")
cat("CRS:", crs(rr), "\n")
cat("Resolution:", paste(res(rr), collapse = " x "), "degrees\n")
cat("Layers (time steps):", nlyr(rr), "\n")
cat("Extent:", paste(ext(rr)[], collapse = ", "), "\n\n")

# Admin boundaries
cat("=== Admin boundaries ===\n")
cat("CRS:", st_crs(admin)$input, "\n")
cat("Features:", nrow(admin), "\n")
cat("Extent:", paste(st_bbox(admin), collapse = ", "), "\n")
```

**What this answers:** "Do my inputs align technically?"

**Output:** a short "data QA" note in your report.

## NUTS boundaries (data_raw)

Read the NUTS regions from GISCO: **NUTS_RG_20M_2024_3035.gpkg** (polygons, 1:20M, EPSG:3035).

```{r read-nuts}
nuts <- read_nuts_raw()
```

### Structure

```{r nuts-structure}
# CRS and extent
st_crs(nuts)
st_bbox(nuts)

# Dimensions and column names
dim(nuts)
names(nuts)
```

### Preview

```{r nuts-preview}
# First rows (geometry omitted for display)
nuts |> 
  st_drop_geometry() |> 
  head(10)
```

### NUTS levels

```{r nuts-levels}
# Typical GISCO attributes: LEVL_CODE (0=country, 1–3=NUTS1–3), NUTS_ID, NAME_LATN, etc.
if ("LEVL_CODE" %in% names(nuts)) {
  nuts |> 
    st_drop_geometry() |> 
    count(LEVL_CODE, name = "n_regions")
}
```

### Simple map (NUTS)

```{r map-nuts, fig.width = 8, fig.height = 6}
library(tmap)
tm_shape(nuts) +
  tm_polygons(col = "grey85", border.col = "grey50") +
  tm_layout(title = "NUTS (2024, 20M)", frame = FALSE)
```

## Next steps

- Filter to study area or NUTS level (e.g. NUTS 1).
- Reproject to WGS84 (EPSG:4326) if needed for E-OBS/crop rasters; save to `data_proc/`.
