---
title: "Main"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
jupyter: R
---

## Setup

Run project setup once (installs packages and downloads `data_raw` from Google Drive if needed). Then load cleaning helpers.

```{r run-setup, include = FALSE}
if (file.exists("R/setup.R")) {
  source("R/setup.R")
}
```

```{r run-download-data_raw, include = FALSE}
source(here::here("R", "download_data_raw.R"))
download_data_raw(root_dir = here::here(), skip_if_exists = TRUE)
```

Load cleaning setup and project packages. Raw data lives in `data_raw/`; we use the NUTS boundaries GeoPackage.

```{r setup, include = FALSE}
source(here::here("R", "data_cleaning.R"))
```

```{r check-terra}
Sys.setenv(PROJ_LIB = system.file("proj", package = "terra"))

library(terra)
```

```{r check-proj}
system.file("proj", package = "terra")
# should be something like ".../R/win-library/4.5/terra/proj"
```

## Step 1 — Load everything and check metadata

### 1.1 Load libraries

```{r load-libs}
library(terra)
library(sf)
library(exactextractr)
library(sfdep)
library(dplyr)
library(tidyr)
library(tmap)
library(here)
```

### 1.2 Read datasets

- E-OBS TG NetCDF (daily)
- E-OBS RR NetCDF (daily)
- SPAM crop GeoTIFF(s)
- Admin boundaries (gpkg/shp)

```{r read-datasets}
# E-OBS temperature (TG) and precipitation (RR)
tg <- rast(here("data_raw/eobs/tg_ens_mean_0.1deg_reg_2011-2022_v27.0e.nc"))
rr <- rast(here("data_raw/eobs/rr_ens_mean_0.1deg_reg_2011-2022_v27.0e.nc"))

# Admin boundaries
admin <- st_read(here("data_raw/admin/NUTS_RG_20M_2024_3035.gpkg"), quiet = TRUE)

# SPAM crop data (adjust path/pattern as needed)
# spam_files <- list.files(here("data_raw/spam"), pattern = "\\.tif$", full.names = TRUE)
# spam <- rast(spam_files)  # or read individual files
```

### 1.3 Inspect CRS + resolution + time dimension

```{r inspect-metadata}
crs(tg); res(tg); nlyr(tg)
crs(rr); res(rr); nlyr(rr)
st_crs(admin)
```

**What this answers:** "Do my inputs align technically?"

**Output:** a short "data QA" note in your report.

## Step 2 — Harmonize study area (DE–BE–NL) + fix geometries

### 2.1 Filter Admin polygons to DE/BE/NL

(Your attribute names differ; adjust `CNTR_CODE` / `ISO` / etc.)

```{r filter-debenl}
# NUTS 2 = “basic regions for regional policies” (most used in EU reporting/policy)
nuts_level <- 2

admin_debenl <- admin %>%
  dplyr::filter(CNTR_CODE %in% c("DE","BE","NL"),
                LEVL_CODE == nuts_level)

if (nrow(admin_debenl) == 0) {
  stop("No admin polygons found for selected countries and NUTS level.")
}
```

### 2.2 Fix geometry + create mask

```{r fix-geometry-mask}
admin_debenl<- st_make_valid(admin_debenl)
mask_union_4326<- st_union(st_transform(admin_debenl,4326))
mask_v<- vect(mask_union_4326)
```

## Step 3 — Clip E-OBS rasters to DE–BE–NL

```{r clip-eobs}
tg_clip<- mask(crop(tg, mask_v), mask_v)
rr_clip<- mask(crop(rr, mask_v), mask_v)
```

**Save Outputs:**

```{r save-clipped-rasters}
dir.create(here("data_proc", "rasters"), showWarnings = FALSE, recursive = TRUE)
# Remove existing files so writeRaster can overwrite (avoids "cannot overwrite" when file is locked)
fn_tg <- here("data_proc/rasters/tg_debenl.tif")
fn_rr <- here("data_proc/rasters/rr_debenl.tif")
if (file.exists(fn_tg)) file.remove(fn_tg)
if (file.exists(fn_rr)) file.remove(fn_rr)
writeRaster(tg_clip, fn_tg, overwrite = TRUE)
writeRaster(rr_clip, fn_rr, overwrite = TRUE)
```

## Step 4 — Select the time window (April–May)

E-OBS NetCDF usually has a time index. Two robust approaches:

### 4.1 If time labels exist (best)

```{r check-time}
head(time(tg_clip), 20)  # first few dates; use time(tg_clip) for full vector
```

If dates exist, filter by month:

```{r filter-april-may}
tg_dates <- as.Date(time(tg_clip))
rr_dates <- as.Date(time(rr_clip))

if (length(tg_dates) != nlyr(tg_clip) || any(is.na(tg_dates))) {
  stop("TG time index is missing or invalid; cannot filter windows.")
}
if (length(rr_dates) != nlyr(rr_clip) || any(is.na(rr_dates))) {
  stop("RR time index is missing or invalid; cannot filter windows.")
}
if (!identical(tg_dates, rr_dates)) {
  stop("TG and RR dates are not aligned. Harmonize time layers before PSI.")
}

sel_am <- which(format(tg_dates, "%m") %in% c("04", "05"))               # main window
sel_mj <- which(format(tg_dates, "%m") %in% c("03", "04", "05", "06"))   # sensitivity
sel_yr <- seq_along(tg_dates)                                             # annual context

if (length(sel_am) == 0) stop("No April-May layers found.")
if (length(sel_mj) == 0) stop("No March-June layers found.")

tg_am <- tg_clip[[sel_am]]
rr_am <- rr_clip[[sel_am]]
```

## Step 5 — Compute daily suitability (PSI) from E-OBS

Use a continuous PSI in [0, 1] (instead of binary 0/1).

Scoring assumptions (piecewise linear):

- Temperature score = 1.0 in 12-25 C, declines to 0 at 8 C and 30 C.
- Rain score = 1.0 for RR <= 1 mm/day, declines to 0 at RR >= 5 mm/day.

### 5.1 Continuous suitability per day (Apr-May)

```{r psi-daily}
compute_daily_psi <- function(tg_stack, rr_stack,
                              t_zero_low = 8, t_opt_low = 12,
                              t_opt_high = 25, t_zero_high = 30,
                              rr_opt = 1, rr_zero = 5) {
  temp_score <- ifel(
    is.na(tg_stack), NA,
    ifel(
      tg_stack <= t_zero_low | tg_stack >= t_zero_high, 0,
      ifel(
        tg_stack >= t_opt_low & tg_stack <= t_opt_high, 1,
        ifel(
          tg_stack < t_opt_low,
          (tg_stack - t_zero_low) / (t_opt_low - t_zero_low),
          (t_zero_high - tg_stack) / (t_zero_high - t_opt_high)
        )
      )
    )
  )

  rain_score <- ifel(
    is.na(rr_stack), NA,
    ifel(
      rr_stack <= rr_opt, 1,
      ifel(
        rr_stack >= rr_zero, 0,
        (rr_zero - rr_stack) / (rr_zero - rr_opt)
      )
    )
  )

  clamp(temp_score * rain_score, lower = 0, upper = 1)
}

psi_daily <- compute_daily_psi(tg_am, rr_am)
time(psi_daily) <- tg_dates[sel_am]
```

### 5.2 Save daily PSI

```{r save-psi-daily}
writeRaster(psi_daily, here("data_proc/rasters/psi_daily.tif"), overwrite=TRUE)
```

**Outputs:** mean, equivalent days, variability.

## Step 6 — Temporal summaries ("spatio-temporal" core)


### 6.1 Mean suitability (mean daily PSI)

```{r psi-mean}
psi_mean<- app(psi_daily, mean, na.rm=TRUE)  # 0–1
```

### 6.2 Equivalent suitable days

```{r psi-days}
sum_or_na <- function(x) {
  if (all(is.na(x))) return(NA_real_)
  sum(x, na.rm = TRUE)
}

psi_days<- app(psi_daily, sum_or_na)  # sum of daily PSI in the window
```

### 6.3 Equivalent deficit days

```{r psi-deficit-days}
psi_deficit_days <- app(1 - psi_daily, sum_or_na)
```

### 6.4 Mean equivalent days per Apr-May season

```{r psi-days-per-year}
n_years_am <- length(unique(format(tg_dates[sel_am], "%Y")))
psi_days_per_year <- psi_days / n_years_am
psi_deficit_days_per_year <- psi_deficit_days / n_years_am
```

### 6.5 Interannual variability (SD across yearly Apr-May means)

```{r psi-sd}
years_am <- unique(format(tg_dates[sel_am], "%Y"))
mean_am_list <- list()
for (y in years_am) {
  idx_y <- which(format(tg_dates[sel_am], "%Y") == y)
  mean_am_list[[y]] <- app(psi_daily[[idx_y]], mean, na.rm = TRUE)
}
psi_interannual_sd <- app(terra::rast(mean_am_list), sd, na.rm = TRUE)
rm(mean_am_list)
gc()
```

Save them:

```{r save-psi-summaries}
writeRaster(psi_mean, here("data_proc/rasters/psi_mean_am.tif"), overwrite=TRUE)
writeRaster(psi_days, here("data_proc/rasters/psi_days_am.tif"), overwrite=TRUE)
writeRaster(psi_deficit_days, here("data_proc/rasters/psi_deficit_days_am.tif"), overwrite=TRUE)
writeRaster(psi_days_per_year, here("data_proc/rasters/psi_days_per_year_am.tif"), overwrite=TRUE)
writeRaster(psi_deficit_days_per_year, here("data_proc/rasters/psi_deficit_days_per_year_am.tif"), overwrite=TRUE)
writeRaster(psi_interannual_sd, here("data_proc/rasters/psi_interannual_sd_am.tif"), overwrite=TRUE)
```

```{r psi-window-context}
# Sensitivity window (Mar-Jun)
tg_mj <- tg_clip[[sel_mj]]
rr_mj <- rr_clip[[sel_mj]]
psi_daily_mj <- compute_daily_psi(tg_mj, rr_mj)
time(psi_daily_mj) <- tg_dates[sel_mj]
psi_mean_mj <- app(psi_daily_mj, mean, na.rm = TRUE)
psi_days_mj <- app(psi_daily_mj, sum_or_na)
psi_deficit_days_mj <- app(1 - psi_daily_mj, sum_or_na)
n_years_mj <- length(unique(format(tg_dates[sel_mj], "%Y")))
psi_days_mj_per_year <- psi_days_mj / n_years_mj
psi_deficit_days_mj_per_year <- psi_deficit_days_mj / n_years_mj

# Annual context window (full year) — computed year-by-year to avoid std::bad_alloc
years_yr <- unique(format(tg_dates[sel_yr], "%Y"))
mean_yr_list <- list()
days_yr_list <- list()
deficit_yr_list <- list()
for (y in years_yr) {
  idx_y <- which(format(tg_dates[sel_yr], "%Y") == y)
  tg_y <- tg_clip[[sel_yr[idx_y]]]
  rr_y <- rr_clip[[sel_yr[idx_y]]]
  psi_y <- compute_daily_psi(tg_y, rr_y)
  mean_yr_list[[y]] <- app(psi_y, mean, na.rm = TRUE)
  days_yr_list[[y]] <- app(psi_y, sum_or_na)
  deficit_yr_list[[y]] <- app(1 - psi_y, sum_or_na)
  rm(tg_y, rr_y, psi_y)
  gc()
}
psi_mean_yr <- mean(terra::rast(mean_yr_list))
psi_days_yr <- sum(terra::rast(days_yr_list))
psi_deficit_days_yr <- sum(terra::rast(deficit_yr_list))
n_years_yr <- length(years_yr)
psi_days_yr_per_year <- psi_days_yr / n_years_yr
psi_deficit_days_yr_per_year <- psi_deficit_days_yr / n_years_yr
rm(mean_yr_list, days_yr_list, deficit_yr_list)
gc()

writeRaster(psi_mean_mj, here("data_proc/rasters/psi_mean_marjun.tif"), overwrite = TRUE)
writeRaster(psi_days_mj, here("data_proc/rasters/psi_days_marjun.tif"), overwrite = TRUE)
writeRaster(psi_deficit_days_mj, here("data_proc/rasters/psi_deficit_days_marjun.tif"), overwrite = TRUE)
writeRaster(psi_days_mj_per_year, here("data_proc/rasters/psi_days_per_year_marjun.tif"), overwrite = TRUE)
writeRaster(psi_deficit_days_mj_per_year, here("data_proc/rasters/psi_deficit_days_per_year_marjun.tif"), overwrite = TRUE)
writeRaster(psi_mean_yr, here("data_proc/rasters/psi_mean_annual.tif"), overwrite = TRUE)
writeRaster(psi_days_yr, here("data_proc/rasters/psi_days_annual.tif"), overwrite = TRUE)
writeRaster(psi_deficit_days_yr, here("data_proc/rasters/psi_deficit_days_annual.tif"), overwrite = TRUE)
writeRaster(psi_days_yr_per_year, here("data_proc/rasters/psi_days_per_year_annual.tif"), overwrite = TRUE)
writeRaster(psi_deficit_days_yr_per_year, here("data_proc/rasters/psi_deficit_days_per_year_annual.tif"), overwrite = TRUE)

window_summary <- tibble::tibble(
  window = c("Apr-May (main)", "Mar-Jun (sensitivity)", "Annual (context)"),
  mean_psi = c(
    as.numeric(global(psi_mean, "mean", na.rm = TRUE)[1, 1]),
    as.numeric(global(psi_mean_mj, "mean", na.rm = TRUE)[1, 1]),
    as.numeric(global(psi_mean_yr, "mean", na.rm = TRUE)[1, 1])
  ),
  equiv_suitable_days = c(
    as.numeric(global(psi_days, "mean", na.rm = TRUE)[1, 1]),
    as.numeric(global(psi_days_mj, "mean", na.rm = TRUE)[1, 1]),
    as.numeric(global(psi_days_yr, "mean", na.rm = TRUE)[1, 1])
  ),
  equiv_deficit_days = c(
    as.numeric(global(psi_deficit_days, "mean", na.rm = TRUE)[1, 1]),
    as.numeric(global(psi_deficit_days_mj, "mean", na.rm = TRUE)[1, 1]),
    as.numeric(global(psi_deficit_days_yr, "mean", na.rm = TRUE)[1, 1])
  ),
  equiv_suitable_days_per_year = c(
    as.numeric(global(psi_days_per_year, "mean", na.rm = TRUE)[1, 1]),
    as.numeric(global(psi_days_mj_per_year, "mean", na.rm = TRUE)[1, 1]),
    as.numeric(global(psi_days_yr_per_year, "mean", na.rm = TRUE)[1, 1])
  ),
  equiv_deficit_days_per_year = c(
    as.numeric(global(psi_deficit_days_per_year, "mean", na.rm = TRUE)[1, 1]),
    as.numeric(global(psi_deficit_days_mj_per_year, "mean", na.rm = TRUE)[1, 1]),
    as.numeric(global(psi_deficit_days_yr_per_year, "mean", na.rm = TRUE)[1, 1])
  ),
  n_days = c(length(sel_am), length(sel_mj), length(sel_yr))
)

dir.create(here("outputs", "tables"), showWarnings = FALSE, recursive = TRUE)
write.csv(window_summary, here("outputs/tables/window_summary.csv"), row.names = FALSE)
window_summary
```

```{r check-rasters}
library(terra)

psi_mean_am <- rast(here("data_proc/rasters/psi_mean_am.tif"))
psi_days_am <- rast(here("data_proc/rasters/psi_days_am.tif"))
psi_deficit_days_am <- rast(here("data_proc/rasters/psi_deficit_days_am.tif"))
psi_days_per_year_am <- rast(here("data_proc/rasters/psi_days_per_year_am.tif"))
psi_interannual_sd_am <- rast(here("data_proc/rasters/psi_interannual_sd_am.tif"))
tg_debenl   <- rast(here("data_proc/rasters/tg_debenl.tif"))
rr_debenl   <- rast(here("data_proc/rasters/rr_debenl.tif"))
psi_daily   <- rast(here("data_proc/rasters/psi_daily.tif"))   # if you saved it too
```

```{r check}
# Are rasters empty?
#global(psi_mean_am, "mean", na.rm=TRUE)
#global(psi_days_am, "max", na.rm=TRUE)

# Value ranges
#global(tg_debenl, range, na.rm=TRUE)
#global(rr_debenl, range, na.rm=TRUE)

# PSI logic check
global(psi_daily, "range", na.rm = TRUE)
```

## Step 7 — Bring SPAM 2020 crop data into the same grid

This is where SPAM shines: it's already ~10 km, but it may still differ from E-OBS grid.

### 7.1 Load SPAM crop raster (example: harvested area for rapeseed)

```{r spam-load}
read_spam_crop <- function(crop_code) {
  candidates <- c(
    sprintf("spam2020_V2r0_global_H_%s_TA.tif", crop_code),
    sprintf("spam2020_V2r0_global_H_%s_A.tif", crop_code)
  )
  found <- candidates[file.exists(here("data_raw/spam", candidates))]
  if (length(found) == 0) {
    stop("Missing SPAM raster for crop code: ", crop_code, ". Expected one of: ", paste(candidates, collapse = ", "))
  }
  if (found[1] != candidates[1]) {
    warning("Using fallback file for ", crop_code, ": ", found[1])
  }
  rast(here("data_raw/spam", found[1]))
}

spam_crops_ha <- list(
  rape = read_spam_crop("RAPE"),
  temf = read_spam_crop("TEMF"),
  sunf = read_spam_crop("SUNF")
)
```

### 7.2 Normalize CRS and resample SPAM to E-OBS template

Use `psi_mean` as the template because it's exactly your analysis grid.

- Do not overwrite CRS labels with `crs(x) <- ...`; that only relabels metadata.
- Use `project()` to reproject and align each crop raster to the PSI template.
- If SPAM is continuous (ha, production): `"bilinear"` is acceptable.

```{r spam-project}
spam_crops_eobs <- lapply(
  spam_crops_ha,
  function(r) project(r, psi_mean, method = "bilinear")
)
```

Save:

```{r spam-save}
writeRaster(spam_crops_eobs$rape, here("data_proc/rasters/spam_rape_ha_eobs.tif"), overwrite=TRUE)
writeRaster(spam_crops_eobs$temf, here("data_proc/rasters/spam_temf_ha_eobs.tif"), overwrite=TRUE)
writeRaster(spam_crops_eobs$sunf, here("data_proc/rasters/spam_sunf_ha_eobs.tif"), overwrite=TRUE)
```

## Step 8 — Convert SPAM “harvested area” to a crop fraction/intensity (optional but strong)

### 8.1 Compute cell area (ha) for the E-OBS grid

```{r cell-area}
cell_area_m2 <- tryCatch(
  cellSize(psi_mean, unit = "m"),
  error = function(e) cellSize(psi_mean)
)
cell_area_ha <- cell_area_m2 / 10000
global(cell_area_m2, "range", na.rm = TRUE)
```

### 8.2 Crop fraction (0–1-ish, can exceed 1 in rare edge cases; clamp)

```{r rape-fraction}
crop_frac <- lapply(spam_crops_eobs, function(r) clamp(r / cell_area_ha, 0, 1))

rape_frac <- crop_frac$rape
temf_frac <- crop_frac$temf
sunf_frac <- crop_frac$sunf

writeRaster(rape_frac, here("data_proc/rasters/rape_fraction.tif"), overwrite=TRUE)
writeRaster(temf_frac, here("data_proc/rasters/temf_fraction.tif"), overwrite=TRUE)
writeRaster(sunf_frac, here("data_proc/rasters/sunf_fraction.tif"), overwrite=TRUE)
```

## Step 9 — Compute crop-weighted suitability and crop-weighted risk

You want metrics that combine climate + crop exposure.

### 9.1 Crop-weighted suitability

```{r rape-suitability-weighted}
crop_suit <- lapply(crop_frac, function(fr) psi_mean * fr)

rape_suit <- crop_suit$rape
temf_suit <- crop_suit$temf
sunf_suit <- crop_suit$sunf

writeRaster(rape_suit, here("data_proc/rasters/rape_suitability_weighted.tif"), overwrite=TRUE)
writeRaster(temf_suit, here("data_proc/rasters/temf_suitability_weighted.tif"), overwrite=TRUE)
writeRaster(sunf_suit, here("data_proc/rasters/sunf_suitability_weighted.tif"), overwrite=TRUE)
```

### 9.2 Crop-weighted deficit exposure (equivalent unsuitable days)

Define daily deficit as (1 - PSI):

```{r rape-bad-exposure}
bad_days <- psi_deficit_days_per_year

crop_bad <- lapply(crop_frac, function(fr) bad_days * fr)

rape_bad_exposure <- crop_bad$rape
temf_bad_exposure <- crop_bad$temf
sunf_bad_exposure <- crop_bad$sunf

writeRaster(rape_bad_exposure, here("data_proc/rasters/rape_bad_exposure.tif"), overwrite=TRUE)
writeRaster(temf_bad_exposure, here("data_proc/rasters/temf_bad_exposure.tif"), overwrite=TRUE)
writeRaster(sunf_bad_exposure, here("data_proc/rasters/sunf_bad_exposure.tif"), overwrite=TRUE)
```

## Step 10 — Aggregate rasters to Admin regions


### 10.1 Prep admin polygons in same CRS as rasters

```{r admin-4326}
admin_4326<- st_transform(admin_debenl, crs(psi_mean))
```

### 10.2 Zonal stats with exactextractr

```{r admin-stats}
admin_stats<- admin_4326%>%
  mutate(
    psi_mean = exact_extract(psi_mean,           ., "mean"),
    psi_days = exact_extract(psi_days,           ., "mean"),
    psi_deficit = exact_extract(psi_deficit_days, ., "mean"),
    psi_days_per_year = exact_extract(psi_days_per_year, ., "mean"),
    psi_deficit_per_year = exact_extract(psi_deficit_days_per_year, ., "mean"),
    psi_interannual_sd = exact_extract(psi_interannual_sd, ., "mean"),
    rape_frac= exact_extract(rape_frac,          ., "mean"),
    rape_suit= exact_extract(rape_suit,          ., "mean"),
    rape_bad = exact_extract(rape_bad_exposure,  ., "mean"),
    temf_frac= exact_extract(temf_frac,          ., "mean"),
    temf_suit= exact_extract(temf_suit,          ., "mean"),
    temf_bad = exact_extract(temf_bad_exposure,  ., "mean"),
    sunf_frac= exact_extract(sunf_frac,          ., "mean"),
    sunf_suit= exact_extract(sunf_suit,          ., "mean"),
    sunf_bad = exact_extract(sunf_bad_exposure,  ., "mean")
  ) %>%
  tidyr::replace_na(list(
    rape_frac = 0, rape_suit = 0, rape_bad = 0,
    temf_frac = 0, temf_suit = 0, temf_bad = 0,
    sunf_frac = 0, sunf_suit = 0, sunf_bad = 0
  )) %>%
  mutate(
    region_id = as.character(NUTS_ID)
  )
```

Save table:

```{r admin-stats-save}
dir.create(here("data_proc", "tables"), showWarnings = FALSE, recursive = TRUE)
write.csv(
  st_drop_geometry(admin_stats),
  here("data_proc/tables/admin_pollination_stats.csv"),
  row.names = FALSE
)
```

**Outputs:**

- A clean regional dataset to map + analyze + rank.

## Step 11 — Spatial weights + Global Moran’s I + LISA (region level)

### 11.1 Prepare neighbors (queen contiguity)

```{r moran-neighbors}
admin_stats_3035<- st_transform(admin_stats, 3035)

nb<- st_contiguity(admin_stats_3035)      # neighbors (queen contiguity)
lw<- st_weights(nb, style="W")           # row-standardized weights
```

### 11.2 Global Moran’s I (example on rape_suit)
```{r}
summary(admin_stats_3035$rape_suit)
table(is.finite(admin_stats_3035$rape_suit))
which(!is.finite(admin_stats_3035$rape_suit))[1:20]  # first few indices
```
```{r global-moran}
library(dplyr)
library(sfdep)

# 1) Keep only regions with valid values
admin_ok <- admin_stats_3035 %>%
  filter(!is.na(rape_suit), is.finite(rape_suit))

# report how many were removed
cat("Rows total:", nrow(admin_stats_3035),
    " | used:", nrow(admin_ok),
    " | removed:", nrow(admin_stats_3035) - nrow(admin_ok), "\n")

# 2) Build neighbors + weights (queen contiguity)
nb_ok <- st_contiguity(admin_ok)
zero_nbr <- lengths(nb_ok) == 0
if (any(zero_nbr)) {
  cat("Dropping isolated regions (0 neighbors):", sum(zero_nbr), "\n")
  admin_ok <- admin_ok[!zero_nbr, ]
  nb_ok <- st_contiguity(admin_ok)
}
lw_ok <- st_weights(nb_ok, style = "W")

# 3) Global Moran's I (sfdep versions without allow_zero argument)
global_m <- global_moran(admin_ok$rape_suit, nb_ok, lw_ok)
global_m

# (optional) permutation-based p-value (stronger for reporting)
global_mc <- global_moran_perm(admin_ok$rape_suit, nb_ok, lw_ok, nsim = 999)
global_mc

```

### 11.3 Cross-crop comparison (exposure + clustering)

```{r crop-compare}
library(dplyr)
library(sfdep)

compare_crops <- c("rape", "temf", "sunf")

exposure_tbl <- tibble(
  crop = compare_crops,
  mean_bad = sapply(compare_crops, function(cr) mean(admin_stats_3035[[paste0(cr, "_bad")]], na.rm = TRUE)),
  p90_bad = sapply(compare_crops, function(cr) as.numeric(quantile(admin_stats_3035[[paste0(cr, "_bad")]], probs = 0.90, na.rm = TRUE))),
  mean_suit = sapply(compare_crops, function(cr) mean(admin_stats_3035[[paste0(cr, "_suit")]], na.rm = TRUE))
) %>%
  arrange(desc(mean_bad))

run_crop_moran <- function(sf_obj, col_name, nsim = 999) {
  sf_ok <- sf_obj %>%
    mutate(value = .data[[col_name]]) %>%
    filter(is.finite(value))

  if (nrow(sf_ok) < 3) {
    return(tibble(moran_I = NA_real_, p_value = NA_real_, n_regions = nrow(sf_ok)))
  }

  nb_ok <- st_contiguity(sf_ok)
  zero_nbr <- lengths(nb_ok) == 0
  if (any(zero_nbr)) {
    sf_ok <- sf_ok[!zero_nbr, ]
    if (nrow(sf_ok) < 3) {
      return(tibble(moran_I = NA_real_, p_value = NA_real_, n_regions = nrow(sf_ok)))
    }
    nb_ok <- st_contiguity(sf_ok)
  }

  lw_ok <- st_weights(nb_ok, style = "W")
  gm <- global_moran(sf_ok$value, nb_ok, lw_ok)
  gm_perm <- global_moran_perm(sf_ok$value, nb_ok, lw_ok, nsim = nsim)

  tibble(
    moran_I = as.numeric(gm$I),
    p_value = as.numeric(gm_perm$p.value),
    n_regions = nrow(sf_ok)
  )
}

cluster_tbl <- bind_rows(lapply(compare_crops, function(cr) {
  out <- run_crop_moran(admin_stats_3035, paste0(cr, "_suit"))
  out$crop <- cr
  out
})) %>%
  select(crop, moran_I, p_value, n_regions) %>%
  arrange(desc(moran_I))

cat("Highest mean exposure (mean *_bad):", exposure_tbl$crop[1], "\n")
cat("Strongest clustering (max Moran's I on *_suit):", cluster_tbl$crop[1], "\n")

dir.create(here("outputs", "tables"), showWarnings = FALSE, recursive = TRUE)
write.csv(exposure_tbl, here("outputs/tables/crop_exposure_comparison.csv"), row.names = FALSE)
write.csv(cluster_tbl, here("outputs/tables/crop_clustering_comparison.csv"), row.names = FALSE)

exposure_tbl
cluster_tbl
```

### 11.4 Local Moran’s I (LISA)

```{r local-moran}
library(dplyr)
library(sfdep)

# 0) Create a stable ID (do this ONCE)
admin_stats_3035 <- admin_stats_3035 %>%
  mutate(
    region_id = if ("region_id" %in% names(.)) as.character(region_id) else if ("NUTS_ID" %in% names(.)) as.character(NUTS_ID) else as.character(row_number())
  )

# 1) Keep only finite values
admin_ok <- admin_stats_3035 %>%
  filter(is.finite(rape_suit))

# 2) Rebuild neighbors/weights
nb_ok <- st_contiguity(admin_ok)
zero_nbr_lisa <- lengths(nb_ok) == 0
if (any(zero_nbr_lisa)) {
  cat("Dropping isolated regions for LISA:", sum(zero_nbr_lisa), "\n")
  admin_ok <- admin_ok[!zero_nbr_lisa, ]
  nb_ok <- st_contiguity(admin_ok)
}
lw_ok <- st_weights(nb_ok, style = "W")

# 3) LISA
lisa_ok <- local_moran(admin_ok$rape_suit, nb_ok, lw_ok)

# 4) Attach results (robust across sfdep output variants)
admin_ok <- admin_ok %>%
  bind_cols(as_tibble(lisa_ok)) %>%
  mutate(
    lisa_i  = ii,
    lisa_p  = p_ii
  )

if ("mean" %in% names(admin_ok)) {
  admin_ok <- admin_ok %>% mutate(lisa_cl = as.character(mean))
} else if ("cluster" %in% names(admin_ok)) {
  admin_ok <- admin_ok %>% mutate(lisa_cl = as.character(cluster))
} else {
  admin_ok <- admin_ok %>% mutate(lisa_cl = NA_character_)
}


```

```{r lisa-join}
# 5) Join back using region_id
admin_stats_3035 <- admin_stats_3035 %>%
  left_join(
    admin_ok %>%
      st_drop_geometry() %>%
      select(region_id, lisa_i, lisa_p, lisa_cl),
    by = "region_id"
  )
```

Verify the join:

```{r lisa-verify}
# Should show NAs for regions dropped from admin_ok (non-finite rape_suit) and the rest filled
table(is.na(admin_stats_3035$lisa_i))
table(admin_stats_3035$lisa_cl, useNA = "ifany")
summary(admin_stats_3035$lisa_p)
```

Create a "significant cluster" label (for mapping):

```{r lisa-sig}
admin_stats_3035 <- admin_stats_3035 %>%
  mutate(
    lisa_sig = case_when(
      is.na(lisa_p) ~ NA_character_,
      lisa_p <= 0.05 ~ as.character(lisa_cl),
      TRUE ~ "Not significant"
    )
  )

table(admin_stats_3035$lisa_sig, useNA = "ifany")
```


Save:

```{r lisa-save}
st_write(
  admin_stats_3035,
  here("data_proc/vectors/admin_stats_lisa.gpkg"),
  delete_dsn = TRUE
)
```

## Step 12 - Create final deliverables (maps, plots, tables)

### 12.1 Maps (minimum set)

```{r maps-deliverables}
library(tmap)
library(here)

dir.create(here("outputs", "maps"), showWarnings = FALSE, recursive = TRUE)
tmap_mode("plot")

save_map <- function(map_obj, filename) {
  tmap_save(map_obj, filename = filename, width = 10, height = 7)
}

admin_4326 <- st_transform(admin_debenl, 4326)

map_psi_mean <- tm_shape(psi_mean) +
  tm_raster(title = "PSI mean") +
  tm_shape(admin_4326) +
  tm_borders(col = "white", lwd = 0.5) +
  tm_layout(title = "Mean PSI (Apr-May)", frame = FALSE)

map_psi_days <- tm_shape(psi_days_per_year) +
  tm_raster(title = "Eq. suitable days/year") +
  tm_shape(admin_4326) +
  tm_borders(col = "white", lwd = 0.5) +
  tm_layout(title = "Mean equivalent suitable days per Apr-May season", frame = FALSE)

map_psi_interannual_sd <- tm_shape(psi_interannual_sd) +
  tm_raster(title = "Interannual PSI SD") +
  tm_shape(admin_4326) +
  tm_borders(col = "white", lwd = 0.5) +
  tm_layout(title = "PSI interannual variability (SD, Apr-May means)", frame = FALSE)

map_rape_frac <- tm_shape(rape_frac) +
  tm_raster(title = "Rape fraction") +
  tm_shape(admin_4326) +
  tm_borders(col = "white", lwd = 0.5) +
  tm_layout(title = "Rapeseed fraction/intensity", frame = FALSE)

map_rape_suit <- tm_shape(rape_suit) +
  tm_raster(title = "Weighted suitability") +
  tm_shape(admin_4326) +
  tm_borders(col = "white", lwd = 0.5) +
  tm_layout(title = "Crop-weighted suitability", frame = FALSE)

save_map(map_psi_mean, here("outputs/maps/map_psi_mean.png"))
save_map(map_psi_days, here("outputs/maps/map_psi_days.png"))
save_map(map_psi_interannual_sd, here("outputs/maps/map_psi_interannual_sd.png"))
save_map(map_rape_frac, here("outputs/maps/map_rape_frac.png"))
save_map(map_rape_suit, here("outputs/maps/map_rape_suit.png"))

if ("lisa_sig" %in% names(admin_stats_3035)) {
  map_lisa <- tm_shape(admin_stats_3035) +
    tm_polygons("lisa_sig", title = "LISA cluster", colorNA = "grey85") +
    tm_layout(title = "LISA clusters (rape_suit)", frame = FALSE)
  save_map(map_lisa, here("outputs/maps/map_lisa_clusters.png"))
}

map_psi_mean
```

### 12.2 Time series plot per region (region x day)

```{r region-day-timeseries}
library(dplyr)
library(ggplot2)
library(here)

dir.create(here("outputs", "plots"), showWarnings = FALSE, recursive = TRUE)
if (!exists("psi_region_day", inherits = FALSE)) {
  psi_region_day <- read.csv(
    here("data_proc", "tables", "psi_region_day.csv"),
    stringsAsFactors = FALSE
  )
}
psi_region_day <- psi_region_day %>%
  mutate(
    region_id = as.character(region_id),
    region_label = as.character(region_label),
    date = as.Date(date),
    psi_daily_mean = as.numeric(psi_daily_mean)
  )

if (!exists("admin_stats_3035", inherits = FALSE)) {
  stats_tbl <- read.csv(
    here("data_proc", "tables", "admin_pollination_stats.csv"),
    stringsAsFactors = FALSE
  )
} else {
  stats_tbl <- if (inherits(admin_stats_3035, "sf")) {
    sf::st_drop_geometry(admin_stats_3035)
  } else {
    admin_stats_3035
  }
}
stats_tbl <- stats_tbl %>%
  mutate(region_id = as.character(region_id))

top_regions <- stats_tbl %>%
  filter(is.finite(rape_bad)) %>%
  arrange(desc(rape_bad)) %>%
  slice_head(n = 12) %>%
  pull(region_id)

plot_df <- psi_region_day %>%
  filter(
    region_id %in% top_regions,
    date >= as.Date("2020-03-01"),
    date <= as.Date("2020-06-30")
  )

if (nrow(plot_df) == 0) {
  stop("No rows after filtering top regions and Mar-Jun 2020 window. Check region_id alignment and date values in psi_region_day.")
}

p_ts <- ggplot(plot_df, aes(x = date, y = psi_daily_mean, color = region_label, group = region_label)) +
  geom_line(alpha = 0.85) +
  labs(
    x = "Date",
    y = "Daily regional mean PSI",
    title = "Daily PSI by region (top 12 by rape_bad) — Mar-Jun 2020"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave(here("outputs/plots/ts_psi_daily_top12_rape_bad_MarJun_2020.png"), p_ts, width = 10, height = 6, dpi = 300)
p_ts

# Second plot: full-year 2020 for top 10 regions by rape_bad
top_regions_10 <- stats_tbl %>%
  filter(is.finite(rape_bad)) %>%
  arrange(desc(rape_bad)) %>%
  slice_head(n = 10) %>%
  pull(region_id)

plot_df_2020 <- psi_region_day %>%
  filter(
    region_id %in% top_regions_10,
    date >= as.Date("2020-01-01"),
    date <= as.Date("2020-12-31")
  )

if (nrow(plot_df_2020) == 0) {
  stop("No rows after filtering top 10 regions and Jan-Dec 2020 window. Check region_id alignment and date values in psi_region_day.")
}

p_ts_2020 <- ggplot(plot_df_2020, aes(x = date, y = psi_daily_mean, color = region_label, group = region_label)) +
  geom_line(alpha = 0.85) +
  labs(
    x = "Date",
    y = "Daily regional mean PSI",
    title = "Daily PSI by region (top 10 by rape_bad) — Jan-Dec 2020"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave(here("outputs/plots/ts_psi_daily_top10_rape_bad_2020.png"), p_ts_2020, width = 10, height = 6, dpi = 300)
p_ts_2020
```

### 12.3 Ranking tables (policy relevance)

```{r ranking-tables}
library(dplyr)
library(here)

dir.create(here("outputs", "tables"), showWarnings = FALSE, recursive = TRUE)

stats_tbl <- read.csv(here("data_proc/tables/admin_pollination_stats.csv"), stringsAsFactors = FALSE)
stats_tbl <- stats_tbl %>%
  mutate(
    region_id = as.character(NUTS_ID)
  )

id_candidates <- c("region_id", "NUTS_ID", "NAME_LATN", "CNTR_CODE", "LEVL_CODE")
id_cols <- id_candidates[id_candidates %in% names(stats_tbl)]

top_rape_bad <- stats_tbl %>%
  filter(is.finite(rape_bad)) %>%
  arrange(desc(rape_bad)) %>%
  slice_head(n = 10)

top_rape_suit <- stats_tbl %>%
  filter(is.finite(rape_suit)) %>%
  arrange(desc(rape_suit)) %>%
  slice_head(n = 10)

top_psi_interannual_sd <- stats_tbl %>%
  filter(is.finite(psi_interannual_sd)) %>%
  arrange(desc(psi_interannual_sd)) %>%
  slice_head(n = 10)

write.csv(top_rape_bad, here("outputs/tables/top10_rape_bad.csv"), row.names = FALSE)
write.csv(top_rape_suit, here("outputs/tables/top10_rape_suit.csv"), row.names = FALSE)
write.csv(top_psi_interannual_sd, here("outputs/tables/top10_psi_interannual_sd.csv"), row.names = FALSE)

top_rape_bad %>% select(any_of(c(id_cols, "rape_bad")))
top_rape_suit %>% select(any_of(c(id_cols, "rape_suit")))
top_psi_interannual_sd %>% select(any_of(c(id_cols, "psi_interannual_sd")))
```
