---
title: "Main"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
jupyter: R
---

## Setup

Run project setup once (installs packages and downloads `data_raw` from Google Drive if needed). Then load cleaning helpers.

```{r run-setup, include = FALSE}
if (file.exists("R/setup.R")) {
  source("R/setup.R")
}
```

```{r run-download-data_raw, include = FALSE}
source(here::here("R", "download_data_raw.R"))
download_data_raw(root_dir = here::here(), skip_if_exists = TRUE)
```

Load cleaning setup and project packages. Raw data lives in `data_raw/`; we use the NUTS boundaries GeoPackage.

```{r setup, include = FALSE}
source(here::here("R", "data_cleaning.R"))
```

```{r check-terra}
Sys.setenv(PROJ_LIB = system.file("proj", package = "terra"))

library(terra)
```

```{r check-proj}
system.file("proj", package = "terra")
# should be something like ".../R/win-library/4.5/terra/proj"
```

## Step 1 — Load everything and check metadata

### 1.1 Load libraries

```{r load-libs}
library(terra)
library(sf)
library(exactextractr)
library(sfdep)
library(dplyr)
library(tidyr)
library(tmap)
library(here)
```

### 1.2 Read datasets

- E-OBS TG NetCDF (daily)
- E-OBS RR NetCDF (daily)
- SPAM crop GeoTIFF(s)
- Admin boundaries (gpkg/shp)

```{r read-datasets}
# E-OBS temperature (TG) and precipitation (RR)
tg <- rast(here("data_raw/eobs/tg_ens_mean_0.1deg_reg_2011-2022_v27.0e.nc"))
rr <- rast(here("data_raw/eobs/rr_ens_mean_0.1deg_reg_2011-2022_v27.0e.nc"))

# Admin boundaries
admin <- st_read(here("data_raw/admin/NUTS_RG_20M_2024_3035.gpkg"), quiet = TRUE)

# SPAM crop data (adjust path/pattern as needed)
# spam_files <- list.files(here("data_raw/spam"), pattern = "\\.tif$", full.names = TRUE)
# spam <- rast(spam_files)  # or read individual files
```

### 1.3 Inspect CRS + resolution + time dimension

```{r inspect-metadata}
crs(tg); res(tg); nlyr(tg)
crs(rr); res(rr); nlyr(rr)
st_crs(admin)
```

**What this answers:** "Do my inputs align technically?"

**Output:** a short "data QA" note in your report.

## Step 2 — Harmonize study area (DE–BE–NL) + fix geometries

### 2.1 Filter Admin polygons to DE/BE/NL

(Your attribute names differ; adjust `CNTR_CODE` / `ISO` / etc.)

```{r filter-debenl}
# NUTS 2 = “basic regions for regional policies” (most used in EU reporting/policy)
nuts_level <- 2

admin_debenl <- admin %>%
  dplyr::filter(CNTR_CODE %in% c("DE","BE","NL"),
                LEVL_CODE == nuts_level)

if (nrow(admin_debenl) == 0) {
  stop("No admin polygons found for selected countries and NUTS level.")
}
```

### 2.2 Fix geometry + create mask

```{r fix-geometry-mask}
admin_debenl<- st_make_valid(admin_debenl)
mask_union_4326<- st_union(st_transform(admin_debenl,4326))
mask_v<- vect(mask_union_4326)
```

## Step 3 — Clip E-OBS rasters to DE–BE–NL

```{r clip-eobs}
tg_clip<- mask(crop(tg, mask_v), mask_v)
rr_clip<- mask(crop(rr, mask_v), mask_v)
```

**Save Outputs:**

```{r save-clipped-rasters}
dir.create(here("data_proc", "rasters"), showWarnings = FALSE, recursive = TRUE)
# Remove existing files so writeRaster can overwrite (avoids "cannot overwrite" when file is locked)
fn_tg <- here("data_proc/rasters/tg_debenl.tif")
fn_rr <- here("data_proc/rasters/rr_debenl.tif")
if (file.exists(fn_tg)) file.remove(fn_tg)
if (file.exists(fn_rr)) file.remove(fn_rr)
writeRaster(tg_clip, fn_tg, overwrite = TRUE)
writeRaster(rr_clip, fn_rr, overwrite = TRUE)
```

## Step 4 — Time index and helper functions

```{r time-index}
tg_dates <- as.Date(time(tg_clip))
rr_dates <- as.Date(time(rr_clip))

if (length(tg_dates) != nlyr(tg_clip) || any(is.na(tg_dates))) {
  stop("TG time index is missing or invalid.")
}
if (length(rr_dates) != nlyr(rr_clip) || any(is.na(rr_dates))) {
  stop("RR time index is missing or invalid.")
}
if (!identical(tg_dates, rr_dates)) {
  stop("TG and RR dates are not aligned.")
}
dates_full <- tg_dates
```

```{r helper-functions}
library(tibble)

select_window_by_mmdd <- function(dates, start_mmdd, end_mmdd) {
  years_chr <- format(dates, "%Y")
  start_dates <- as.Date(paste0(years_chr, "-", start_mmdd))
  end_dates <- as.Date(paste0(years_chr, "-", end_mmdd))
  as.logical(!is.na(start_dates) & !is.na(end_dates) & dates >= start_dates & dates <= end_dates)
}

sum_or_na <- function(x) {
  if (all(is.na(x))) return(NA_real_)
  sum(x, na.rm = TRUE)
}

summarize_window <- function(psi_daily_full, dates_full, start_mmdd, end_mmdd) {
  sel <- select_window_by_mmdd(dates_full, start_mmdd, end_mmdd)
  if (!any(sel)) {
    stop("No layers selected for window ", start_mmdd, " to ", end_mmdd)
  }

  psi_w <- psi_daily_full[[which(sel)]]
  dates_w <- dates_full[sel]
  years_w <- unique(format(dates_w, "%Y"))

  mean_r <- app(psi_w, mean, na.rm = TRUE)
  equiv_days_r <- app(psi_w, sum_or_na)
  deficit_days_r <- app(1 - psi_w, sum_or_na)
  equiv_days_per_year_r <- equiv_days_r / length(years_w)
  deficit_days_per_year_r <- deficit_days_r / length(years_w)

  yearly_means <- list()
  for (yy in years_w) {
    idx_y <- which(format(dates_w, "%Y") == yy)
    yearly_means[[yy]] <- app(psi_w[[idx_y]], mean, na.rm = TRUE)
  }
  interannual_sd_r <- app(terra::rast(yearly_means), sd, na.rm = TRUE)

  list(
    mean = mean_r,
    equiv_days = equiv_days_r,
    equiv_days_per_year = equiv_days_per_year_r,
    deficit_days = deficit_days_r,
    deficit_days_per_year = deficit_days_per_year_r,
    interannual_sd = interannual_sd_r,
    dates = dates_w,
    n_days = length(dates_w),
    n_years = length(years_w)
  )
}

make_crop_fraction <- function(spam_raster, template_raster) {
  spam_aligned <- project(spam_raster, template_raster, method = "bilinear")
  cell_area_m2 <- tryCatch(
    cellSize(template_raster, unit = "m"),
    error = function(e) cellSize(template_raster)
  )
  cell_area_ha <- cell_area_m2 / 10000
  clamp(spam_aligned / cell_area_ha, 0, 1)
}

plot_and_save_raster <- function(r, title_txt, filename, admin_borders, admin_lvl0_borders) {
  m <- tm_shape(r) +
    tm_raster(title = title_txt) +
    tm_shape(admin_borders) +
    tm_borders(col = "white", lwd = 0.5) +
    tm_shape(admin_lvl0_borders) +
    tm_borders(col = "black", lwd = 1.2) +
    tm_layout(title = title_txt, frame = FALSE)
  tmap_save(m, filename = filename, width = 10, height = 7)
  invisible(m)
}

build_region_day_table <- function(psi_stack, dates_stack, admin_sf) {
  admin_ts <- sf::st_transform(admin_sf, terra::crs(psi_stack))
  admin_labels <- sf::st_drop_geometry(admin_ts) %>%
    mutate(region_id = as.character(NUTS_ID))

  label_candidates <- c("NAME_LATN", "NUTS_NAME", "NAME_ENGL")
  label_col <- label_candidates[label_candidates %in% names(admin_labels)]
  admin_labels$region_label <- if (length(label_col) > 0) as.character(admin_labels[[label_col[1]]]) else admin_labels$region_id

  daily_wide <- exactextractr::exact_extract(psi_stack, admin_ts, "mean") %>% tibble::as_tibble()
  if (nrow(daily_wide) != nrow(admin_labels)) {
    stop("Row mismatch while extracting daily regional PSI.")
  }
  daily_wide$region_id <- admin_labels$region_id

  layer_lookup <- tibble(
    layer = paste0("mean.", names(psi_stack)),
    layer_index = seq_len(terra::nlyr(psi_stack)),
    date = as.Date(dates_stack)
  )

  daily_wide %>%
    tidyr::pivot_longer(-region_id, names_to = "layer", values_to = "psi_daily_mean") %>%
    left_join(layer_lookup, by = "layer") %>%
    left_join(admin_labels %>% select(region_id, region_label), by = "region_id")
}
```

## Step 5 — Compute daily PSI once (full period)

Continuous PSI (0-1), then reused for baseline and crop-specific windows.

```{r psi-daily-full}
compute_daily_psi <- function(tg_stack, rr_stack,
                              t_zero_low = 8, t_opt_low = 12,
                              t_opt_high = 25, t_zero_high = 30,
                              rr_opt = 1, rr_zero = 5) {
  temp_score <- ifel(
    is.na(tg_stack), NA,
    ifel(
      tg_stack <= t_zero_low | tg_stack >= t_zero_high, 0,
      ifel(
        tg_stack >= t_opt_low & tg_stack <= t_opt_high, 1,
        ifel(
          tg_stack < t_opt_low,
          (tg_stack - t_zero_low) / (t_opt_low - t_zero_low),
          (t_zero_high - tg_stack) / (t_zero_high - t_opt_high)
        )
      )
    )
  )

  rain_score <- ifel(
    is.na(rr_stack), NA,
    ifel(rr_stack <= rr_opt, 1, ifel(rr_stack >= rr_zero, 0, (rr_zero - rr_stack) / (rr_zero - rr_opt)))
  )
  clamp(temp_score * rain_score, lower = 0, upper = 1)
}

psi_daily_full <- compute_daily_psi(tg_clip, rr_clip)
time(psi_daily_full) <- dates_full

dir.create(here("data_proc", "rasters"), showWarnings = FALSE, recursive = TRUE)
writeRaster(psi_daily_full, here("data_proc/rasters/psi_daily_full.tif"), overwrite = TRUE)
```

## Step 6 — Baseline window (Apr-May) + crop-specific windows

Equivalent suitable days are continuous sums of daily PSI, not binary day counts.

```{r window-definitions}
crop_windows <- tibble(
  crop = c("rape", "temf", "sunf"),
  start_mmdd = c("04-01", "04-05", "07-10"),
  end_mmdd   = c("05-31", "05-10", "08-10")
)

baseline_window <- list(start_mmdd = "04-01", end_mmdd = "05-31")
```

```{r summarize-windows}
# Baseline (kept for comparability)
baseline_summary <- summarize_window(
  psi_daily_full = psi_daily_full,
  dates_full = dates_full,
  start_mmdd = baseline_window$start_mmdd,
  end_mmdd = baseline_window$end_mmdd
)
psi_mean <- baseline_summary$mean
psi_days_per_year <- baseline_summary$equiv_days_per_year
psi_interannual_sd <- baseline_summary$interannual_sd

writeRaster(baseline_summary$mean, here("data_proc/rasters/psi_mean_am.tif"), overwrite = TRUE)
writeRaster(baseline_summary$equiv_days, here("data_proc/rasters/psi_equiv_days_am.tif"), overwrite = TRUE)
writeRaster(baseline_summary$equiv_days_per_year, here("data_proc/rasters/psi_equiv_days_per_year_am.tif"), overwrite = TRUE)
writeRaster(baseline_summary$deficit_days, here("data_proc/rasters/psi_deficit_days_am.tif"), overwrite = TRUE)
writeRaster(baseline_summary$deficit_days_per_year, here("data_proc/rasters/psi_deficit_days_per_year_am.tif"), overwrite = TRUE)
writeRaster(baseline_summary$interannual_sd, here("data_proc/rasters/psi_interannual_sd_am.tif"), overwrite = TRUE)

crop_psi <- list()
for (i in seq_len(nrow(crop_windows))) {
  cr <- crop_windows$crop[i]
  st_mmdd <- crop_windows$start_mmdd[i]
  en_mmdd <- crop_windows$end_mmdd[i]
  crop_psi[[cr]] <- summarize_window(psi_daily_full, dates_full, st_mmdd, en_mmdd)

  writeRaster(crop_psi[[cr]]$mean, here("data_proc/rasters", sprintf("psi_mean_%s.tif", cr)), overwrite = TRUE)
  writeRaster(crop_psi[[cr]]$equiv_days, here("data_proc/rasters", sprintf("psi_equiv_days_%s.tif", cr)), overwrite = TRUE)
  writeRaster(crop_psi[[cr]]$equiv_days_per_year, here("data_proc/rasters", sprintf("psi_equiv_days_per_year_%s.tif", cr)), overwrite = TRUE)
  writeRaster(crop_psi[[cr]]$deficit_days, here("data_proc/rasters", sprintf("psi_deficit_days_%s.tif", cr)), overwrite = TRUE)
  writeRaster(crop_psi[[cr]]$deficit_days_per_year, here("data_proc/rasters", sprintf("psi_deficit_days_per_year_%s.tif", cr)), overwrite = TRUE)
  writeRaster(crop_psi[[cr]]$interannual_sd, here("data_proc/rasters", sprintf("psi_interannual_sd_%s.tif", cr)), overwrite = TRUE)
}
```

## Step 7 — SPAM crops and crop fraction/intensity

```{r spam-load-and-fraction}
read_spam_crop <- function(crop_code) {
  candidates <- c(
    sprintf("spam2020_V2r0_global_H_%s_TA.tif", crop_code),
    sprintf("spam2020_V2r0_global_H_%s_A.tif", crop_code)
  )
  found <- candidates[file.exists(here("data_raw/spam", candidates))]
  if (length(found) == 0) {
    stop("Missing SPAM raster for crop code: ", crop_code, ". Expected one of: ", paste(candidates, collapse = ", "))
  }
  rast(here("data_raw/spam", found[1]))
}

spam_crops_ha <- list(
  rape = read_spam_crop("RAPE"),
  temf = read_spam_crop("TEMF"),
  sunf = read_spam_crop("SUNF")
)

template_r <- baseline_summary$mean
crop_frac <- list()
for (cr in names(spam_crops_ha)) {
  crop_frac[[cr]] <- make_crop_fraction(spam_crops_ha[[cr]], template_r)
  writeRaster(crop_frac[[cr]], here("data_proc/rasters", sprintf("crop_frac_%s.tif", cr)), overwrite = TRUE)
}
```

## Step 8 — Crop-weighted suitability and deficit exposure

`crop_suit_{crop} = psi_mean_{crop} * crop_frac_{crop}` is fraction-based.  
`crop_bad_{crop} = psi_deficit_days_per_year_{crop} * crop_frac_{crop}` is day-based.  
These are complementary indicators and are not directly comparable by magnitude.

```{r crop-weighted}
crop_suit <- list()
crop_bad <- list()
for (cr in names(crop_frac)) {
  crop_suit[[cr]] <- crop_psi[[cr]]$mean * crop_frac[[cr]]
  crop_bad[[cr]] <- crop_psi[[cr]]$deficit_days_per_year * crop_frac[[cr]]

  writeRaster(crop_suit[[cr]], here("data_proc/rasters", sprintf("crop_suit_%s.tif", cr)), overwrite = TRUE)
  writeRaster(crop_bad[[cr]], here("data_proc/rasters", sprintf("crop_bad_%s.tif", cr)), overwrite = TRUE)
}
```

## Step 9 — Aggregate to admin regions (NUTS2, stable region_id)

```{r admin-stats}
admin_4326 <- st_transform(admin_debenl, crs(template_r))
admin_stats <- admin_4326 %>%
  mutate(
    region_id = as.character(NUTS_ID),
    psi_mean_am = exact_extract(baseline_summary$mean, ., "mean"),
    psi_equiv_days_per_year_am = exact_extract(baseline_summary$equiv_days_per_year, ., "mean"),
    psi_interannual_sd_am = exact_extract(baseline_summary$interannual_sd, ., "mean")
  )

for (cr in crop_windows$crop) {
  admin_stats[[paste0("psi_mean_", cr)]] <- exact_extract(crop_psi[[cr]]$mean, admin_stats, "mean")
  admin_stats[[paste0("psi_equiv_days_", cr)]] <- exact_extract(crop_psi[[cr]]$equiv_days, admin_stats, "mean")
  admin_stats[[paste0("psi_equiv_days_per_year_", cr)]] <- exact_extract(crop_psi[[cr]]$equiv_days_per_year, admin_stats, "mean")
  admin_stats[[paste0("psi_deficit_days_", cr)]] <- exact_extract(crop_psi[[cr]]$deficit_days, admin_stats, "mean")
  admin_stats[[paste0("psi_deficit_days_per_year_", cr)]] <- exact_extract(crop_psi[[cr]]$deficit_days_per_year, admin_stats, "mean")
  admin_stats[[paste0("psi_interannual_sd_", cr)]] <- exact_extract(crop_psi[[cr]]$interannual_sd, admin_stats, "mean")
  admin_stats[[paste0("crop_frac_", cr)]] <- exact_extract(crop_frac[[cr]], admin_stats, "mean")
  admin_stats[[paste0("crop_suit_", cr)]] <- exact_extract(crop_suit[[cr]], admin_stats, "mean")
  admin_stats[[paste0("crop_bad_", cr)]] <- exact_extract(crop_bad[[cr]], admin_stats, "mean")
}

crop_cols <- grep("^(crop_frac_|crop_suit_|crop_bad_)", names(admin_stats), value = TRUE)
for (cc in crop_cols) {
  admin_stats[[cc]][is.na(admin_stats[[cc]])] <- 0
}
```

```{r admin-stats-save}
dir.create(here("data_proc", "tables"), showWarnings = FALSE, recursive = TRUE)
write.csv(st_drop_geometry(admin_stats), here("data_proc/tables/admin_pollination_stats.csv"), row.names = FALSE)
```

## Step 10 — Global Moran’s I + Local Moran (LISA) per crop metric

```{r spatial-stats}
library(sfdep)

admin_stats_3035 <- st_transform(admin_stats, 3035)
compare_crops <- crop_windows$crop
compare_metrics <- c("crop_bad", "crop_suit")

run_global_moran <- function(sf_obj, value_col, nsim = 999) {
  sf_ok <- sf_obj %>%
    mutate(value = .data[[value_col]]) %>%
    filter(is.finite(value))
  if (nrow(sf_ok) < 3) return(tibble(moran_i = NA_real_, p_value = NA_real_, n_regions = nrow(sf_ok)))

  nb_ok <- st_contiguity(sf_ok)
  keep <- lengths(nb_ok) > 0
  sf_ok <- sf_ok[keep, ]
  if (nrow(sf_ok) < 3) return(tibble(moran_i = NA_real_, p_value = NA_real_, n_regions = nrow(sf_ok)))
  nb_ok <- st_contiguity(sf_ok)
  lw_ok <- st_weights(nb_ok, style = "W")

  gm <- global_moran(sf_ok$value, nb_ok, lw_ok)
  gp <- global_moran_perm(sf_ok$value, nb_ok, lw_ok, nsim = nsim)
  tibble(moran_i = as.numeric(gm$I), p_value = as.numeric(gp$p.value), n_regions = nrow(sf_ok))
}

clustering_tbl <- bind_rows(lapply(compare_crops, function(cr) {
  bind_rows(lapply(compare_metrics, function(mt) {
    vcol <- paste0(mt, "_", cr)
    out <- run_global_moran(admin_stats_3035, vcol)
    out$crop <- cr
    out$metric <- mt
    out
  }))
})) %>%
  select(crop, metric, moran_i, p_value, n_regions)

for (cr in compare_crops) {
  val_col <- paste0("crop_bad_", cr)
  sf_ok <- admin_stats_3035 %>%
    mutate(value = .data[[val_col]]) %>%
    filter(is.finite(value))
  nb_ok <- st_contiguity(sf_ok)
  keep <- lengths(nb_ok) > 0
  sf_ok <- sf_ok[keep, ]
  nb_ok <- st_contiguity(sf_ok)
  lw_ok <- st_weights(nb_ok, style = "W")
  lisa <- local_moran(sf_ok$value, nb_ok, lw_ok)
  lisa_tbl <- as_tibble(lisa)

  sf_ok <- sf_ok %>%
    bind_cols(lisa_tbl) %>%
    mutate(
      tmp_i = ii,
      tmp_p = p_ii
    )

  if ("mean" %in% names(sf_ok)) {
    sf_ok <- sf_ok %>% mutate(tmp_cl = as.character(mean))
  } else if ("cluster" %in% names(sf_ok)) {
    sf_ok <- sf_ok %>% mutate(tmp_cl = as.character(cluster))
  } else {
    sf_ok <- sf_ok %>% mutate(tmp_cl = NA_character_)
  }

  sf_ok <- sf_ok %>%
    mutate(
      tmp_sig = case_when(
        is.na(tmp_p) ~ NA_character_,
        tmp_p <= 0.05 ~ tmp_cl,
        TRUE ~ "Not significant"
      )
    ) %>%
    st_drop_geometry() %>%
    select(region_id, tmp_i, tmp_p, tmp_cl, tmp_sig)

  admin_stats_3035 <- admin_stats_3035 %>%
    left_join(sf_ok, by = "region_id") %>%
    rename(
      !!paste0("lisa_i_crop_bad_", cr) := tmp_i,
      !!paste0("lisa_p_crop_bad_", cr) := tmp_p,
      !!paste0("lisa_cl_crop_bad_", cr) := tmp_cl,
      !!paste0("lisa_sig_crop_bad_", cr) := tmp_sig
    )
}

dir.create(here("data_proc", "vectors"), showWarnings = FALSE, recursive = TRUE)
st_write(admin_stats_3035, here("data_proc/vectors/admin_stats_lisa.gpkg"), delete_layer = TRUE)

write.csv(clustering_tbl, here("data_proc/tables/crop_clustering_comparison.csv"), row.names = FALSE)
```

## Step 11 — Maps

```{r maps-deliverables}
library(tmap)

dir.create(here("outputs", "maps"), showWarnings = FALSE, recursive = TRUE)
tmap_mode("plot")

admin_borders <- st_transform(admin_debenl, 4326)
admin_lvl0_borders <- admin_debenl %>%
  group_by(CNTR_CODE) %>%
  summarise(do_union = TRUE, .groups = "drop") %>%
  st_make_valid() %>%
  st_transform(4326)

# Baseline maps
plot_and_save_raster(baseline_summary$mean, "Mean PSI (Apr-May baseline)", here("outputs/maps/map_psi_mean.png"), admin_borders, admin_lvl0_borders)
plot_and_save_raster(baseline_summary$equiv_days_per_year, "Equivalent suitable days/year (Apr-May baseline)", here("outputs/maps/map_psi_days.png"), admin_borders, admin_lvl0_borders)
plot_and_save_raster(baseline_summary$interannual_sd, "Interannual PSI SD (Apr-May baseline)", here("outputs/maps/map_psi_interannual_sd.png"), admin_borders, admin_lvl0_borders)

# Crop-specific maps
for (cr in compare_crops) {
  plot_and_save_raster(crop_psi[[cr]]$mean, sprintf("PSI mean (%s window)", cr), here("outputs/maps", sprintf("map_psi_mean_%s.png", cr)), admin_borders, admin_lvl0_borders)
  plot_and_save_raster(crop_psi[[cr]]$equiv_days_per_year, sprintf("Equivalent suitable days/year (%s window)", cr), here("outputs/maps", sprintf("map_psi_equiv_days_per_year_%s.png", cr)), admin_borders, admin_lvl0_borders)
  plot_and_save_raster(crop_frac[[cr]], sprintf("Crop fraction (%s)", cr), here("outputs/maps", sprintf("map_crop_frac_%s.png", cr)), admin_borders, admin_lvl0_borders)
  plot_and_save_raster(crop_suit[[cr]], sprintf("Crop-weighted suitability (%s)", cr), here("outputs/maps", sprintf("map_crop_suit_%s.png", cr)), admin_borders, admin_lvl0_borders)
  plot_and_save_raster(crop_bad[[cr]], sprintf("Crop-weighted deficit exposure (%s)", cr), here("outputs/maps", sprintf("map_crop_bad_%s.png", cr)), admin_borders, admin_lvl0_borders)

  lisa_col <- paste0("lisa_sig_crop_bad_", cr)
  if (lisa_col %in% names(admin_stats_3035)) {
    map_lisa <- tm_shape(admin_stats_3035) +
      tm_polygons(lisa_col, title = "LISA cluster", colorNA = "grey85") +
      tm_layout(title = sprintf("LISA clusters (%s crop_bad)", cr), frame = FALSE)
    tmap_save(map_lisa, filename = here("outputs/maps", sprintf("map_lisa_%s_bad.png", cr)), width = 10, height = 7)
  }
}
```

## Step 12 — Time-series plots for 2020 crop windows (top 12 by crop_bad)

```{r region-day-timeseries}
library(ggplot2)

dir.create(here("outputs", "plots"), showWarnings = FALSE, recursive = TRUE)

admin_stats_tbl <- st_drop_geometry(admin_stats_3035)

for (i in seq_len(nrow(crop_windows))) {
  cr <- crop_windows$crop[i]
  st_mmdd <- crop_windows$start_mmdd[i]
  en_mmdd <- crop_windows$end_mmdd[i]

  sel_crop <- select_window_by_mmdd(dates_full, st_mmdd, en_mmdd)
  dates_crop <- dates_full[sel_crop]
  psi_crop_daily <- psi_daily_full[[which(sel_crop)]]
  time(psi_crop_daily) <- dates_crop

  psi_region_day <- build_region_day_table(psi_crop_daily, dates_crop, admin_debenl) %>%
    mutate(
      region_id = as.character(region_id),
      region_label = as.character(region_label),
      date = as.Date(date),
      psi_daily_mean = as.numeric(psi_daily_mean)
    )

  write.csv(psi_region_day, here("data_proc/tables", sprintf("psi_region_day_%s.csv", cr)), row.names = FALSE)

  top_regions <- admin_stats_tbl %>%
    filter(is.finite(.data[[paste0("crop_bad_", cr)]])) %>%
    arrange(desc(.data[[paste0("crop_bad_", cr)]])) %>%
    slice_head(n = 12) %>%
    pull(region_id)

  plot_df <- psi_region_day %>%
    filter(
      region_id %in% top_regions,
      date >= as.Date(paste0("2020-", st_mmdd)),
      date <= as.Date(paste0("2020-", en_mmdd))
    )

  if (nrow(plot_df) == 0) {
    stop("No rows for ", cr, " after filtering 2020 crop window.")
  }

  p_ts <- ggplot(plot_df, aes(x = date, y = psi_daily_mean, color = region_label, group = region_label)) +
    geom_line(alpha = 0.85) +
    labs(
      x = "Date",
      y = "Daily regional mean PSI",
      title = sprintf("Daily PSI by region (top 12 by %s bad) — %s to %s, 2020", cr, st_mmdd, en_mmdd)
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")

  ggsave(
    here("outputs/plots", sprintf("ts_psi_daily_top12_%s_bad_%s_%s_2020.png", cr, gsub("-", "", st_mmdd), gsub("-", "", en_mmdd))),
    p_ts, width = 10, height = 6, dpi = 300
  )
}
```

## Step 13 — Comparison and ranking tables

```{r ranking-tables}
dir.create(here("outputs", "tables"), showWarnings = FALSE, recursive = TRUE)

admin_stats_tbl <- st_drop_geometry(admin_stats_3035)

ranking_rows <- list()
for (cr in compare_crops) {
  bad_col <- paste0("crop_bad_", cr)
  suit_col <- paste0("crop_suit_", cr)

  top_bad <- admin_stats_tbl %>%
    filter(is.finite(.data[[bad_col]])) %>%
    arrange(desc(.data[[bad_col]])) %>%
    slice_head(n = 10) %>%
    mutate(crop = cr, metric = "crop_bad", rank = row_number(), value = .data[[bad_col]])

  top_suit <- admin_stats_tbl %>%
    filter(is.finite(.data[[suit_col]])) %>%
    arrange(desc(.data[[suit_col]])) %>%
    slice_head(n = 10) %>%
    mutate(crop = cr, metric = "crop_suit", rank = row_number(), value = .data[[suit_col]])

  write.csv(top_bad, here("outputs/tables", sprintf("top10_%s_bad.csv", cr)), row.names = FALSE)
  write.csv(top_suit, here("outputs/tables", sprintf("top10_%s_suit.csv", cr)), row.names = FALSE)

  ranking_rows[[paste0(cr, "_bad")]] <- top_bad %>% select(crop, metric, rank, region_id, NUTS_ID, NAME_LATN, value)
  ranking_rows[[paste0(cr, "_suit")]] <- top_suit %>% select(crop, metric, rank, region_id, NUTS_ID, NAME_LATN, value)
}

crop_exposure_comparison <- bind_rows(ranking_rows)
write.csv(crop_exposure_comparison, here("data_proc/tables/crop_exposure_comparison.csv"), row.names = FALSE)
write.csv(crop_exposure_comparison, here("outputs/tables/crop_exposure_comparison.csv"), row.names = FALSE)
```
